---
title: "Cluster-Relative Mean Reversion on S&P 500 Sector ETFs"
subtitle: "FIN452 — Quantitative Trading Final Project"
author: "Will Thompson"
date: "`r Sys.Date()`"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    toc-float: true
    theme: flatly
    fig-width: 10
    fig-height: 6
execute:
  warning: false
  message: false
  cache: true
  echo: false
---

```{r setup}
#| cache: false
#| include: false

# ── Core tidyverse ─────────────────────────────────────────────────────────────
library(tidyverse)
library(lubridate)

# ── Market data ────────────────────────────────────────────────────────────────
library(tidyquant)
library(timetk)
library(quantmod)

# ── Rolling statistics ─────────────────────────────────────────────────────────
library(slider)
library(TTR)

# ── Tidymodels (preprocessing + data splitting) ────────────────────────────────
library(tidymodels)
library(rsample)
library(recipes)

# ── Clustering ─────────────────────────────────────────────────────────────────
library(cluster)      # silhouette()
library(factoextra)   # fviz_dend(), fviz_silhouette()

# ── Performance analytics ──────────────────────────────────────────────────────
library(PerformanceAnalytics)   # SharpeRatio, Omega, chart.Drawdown, table.Drawdowns


# ── Visualization ──────────────────────────────────────────────────────────────
library(patchwork)
library(scales)
library(viridis)
library(plotly)

# ── Tables ─────────────────────────────────────────────────────────────────────
library(gt)
library(knitr)

# Resolve namespace conflicts
tidymodels_prefer()
```

```{r parameters}
#| cache: false
#| include: false

# ── Date ranges ────────────────────────────────────────────────────────────────
DATA_START  <- as.Date("2000-01-01")   # Extra history for rolling-feature warmup
TRAIN_START <- as.Date("2000-06-01")   # First full month of training period
TRAIN_END   <- as.Date("2024-12-31")   # End of training period
TEST_START  <- as.Date("2025-01-01")   # Out-of-sample begins
TEST_END    <- as.Date("2026-03-31")   # Inclusive through March 2026

# ── Feature engineering windows (trading days) ─────────────────────────────────
VOL_SHORT   <- 20L   # ~1-month rolling volatility            [DEFAULT: 20]
VOL_LONG    <- 60L   # ~3-month rolling volatility            [DEFAULT: 60]
CORR_WINDOW <- 60L   # Rolling correlation / beta vs SPY      [DEFAULT: 60]
RET_WINDOW  <- 60L   # Rolling cumulative return for residual [DEFAULT: 60]

# ── Clustering ─────────────────────────────────────────────────────────────────
# N_CLUSTERS: Chosen using elbow method (WSS) in section 6.
# With 10-11 ETFs, k=3 gives ~3-4 members per cluster on average.
# [UNRESOLVED] Validate: run elbow method and silhouette across k = 2..5.
N_CLUSTERS  <- 3L
LINK_METHOD <- "ward.D2"   # Ward linkage: minimises within-cluster variance (preferred for finance)

# Features used for clustering
CLUSTER_FEATURES <- c("vol_20", "vol_60", "corr_spy_60", "beta_spy_60", "rel_ret_60")

# ── Kalman filter ──────────────────────────────────────────────────────────────
# [UNRESOLVED] Q/R ratio controls smoothing: low Q/R = slow adaptation (smoother).
# Default Q/R = 0.1 (moderately smooth). Sensitivity tested in section 13.
KF_Q <- 0.001   # Process noise variance  [DEFAULT: 0.001]
KF_R <- 0.01    # Observation noise variance [DEFAULT: 0.01]

# ── Signal thresholds ──────────────────────────────────────────────────────────
# [UNRESOLVED] Margin of safety: only trade |z| >= this value.
SIGNAL_THRESHOLD <- 1.0   # [DEFAULT: 1.0 ~ 1 standard deviation]

# ── Portfolio construction ─────────────────────────────────────────────────────
# [UNRESOLVED] Balances diversification vs signal dilution.
MAX_LONGS  <- 3L   # Maximum long positions per rebalance  [DEFAULT: 3]
MAX_SHORTS <- 3L   # Maximum short positions per rebalance [DEFAULT: 3]

# ── Execution assumption (fixed) ───────────────────────────────────────────────
# Signal at month-end close → position return = next month close-to-close.
# Strictly causal: no future information used at signal generation time.
```

```{r universe}
#| cache: false
#| include: false

TICKERS <- c(
  "SPY",   # Benchmark — S&P 500
  "XLB",   # Materials
  "XLC",   # Communication Services (from June 2018)
  "XLE",   # Energy
  "XLF",   # Financials
  "XLI",   # Industrials
  "XLK",   # Technology
  "XLP",   # Consumer Staples
  "XLRE",  # Real Estate (from October 2015)
  "XLV",   # Health Care
  "XLU",   # Utilities
  "XLY"    # Consumer Discretionary
)

SECTOR_TICKERS <- setdiff(TICKERS, "SPY")
cat("Universe: SPY (benchmark) +", length(SECTOR_TICKERS), "sector ETFs\n")
```

```{r data-download}
#| include: false

# tq_get() returns: symbol, date, open, high, low, close, volume, adjusted
# 'adjusted' = split and dividend adjusted close — used for all return calculations

raw_prices <- tq_get(
  TICKERS,
  from = DATA_START,
  to   = TEST_END
)

cat("Download complete.\n")
cat("Date range  :", format(min(raw_prices$date)), "to", format(max(raw_prices$date)), "\n")
cat("Rows        :", nrow(raw_prices), "| Symbols:", n_distinct(raw_prices$symbol), "\n")
```

```{r preprocessing}
#| include: false

# Keep adjusted close for return calculations.
# Mirror course notes: retClCl (close-to-close), retOpCl (open-to-close same day),
# retClOp (open vs prior close). Used for precise execution assumptions.

prices <- raw_prices |>
  select(symbol, date, open, close, adjusted) |>
  arrange(symbol, date)

returns_daily <- prices |>
  group_by(symbol) |>
  arrange(date) |>
  mutate(
    retClCl = adjusted / dplyr::lag(adjusted) - 1,    # Close-to-close (hold overnight)
    retOpCl = (close - open) / close,                 # Open-to-close same day (intraday)
    retClOp = open / dplyr::lag(adjusted) - 1         # Prior close to open (gap return)
  ) |>
  ungroup() |>
  filter(!is.na(retClCl))

cat("Daily returns computed.\n")
cat("Total rows:", nrow(returns_daily), "\n")
```

```{r alignment}
#| include: false

# Align all series to the SPY trading calendar (removes non-trading days).
spy_calendar <- returns_daily |>
  filter(symbol == "SPY") |>
  pull(date)

returns_daily <- returns_daily |>
  filter(date %in% spy_calendar)

spy_rets_wide <- returns_daily |>
  filter(symbol == "SPY") |>
  select(date, SPY = retClCl)

cat("Calendar aligned to SPY. Trading days:", n_distinct(returns_daily$date), "\n")
```

```{r train-test-split}
#| include: false

# Strict temporal split — no testing period data used during model design.
returns_train <- returns_daily |> filter(date >= TRAIN_START, date <= TRAIN_END)
returns_test  <- returns_daily |> filter(date >= TEST_START,  date <= TEST_END)

# Leakage guard
overlap <- intersect(returns_train$date, returns_test$date)
stopifnot("LEAKAGE DETECTED: train and test dates overlap!" = length(overlap) == 0)

cat("Training:", format(TRAIN_START), "to", format(TRAIN_END),
    "—", n_distinct(returns_train$date), "trading days\n")
cat("Testing :", format(TEST_START),  "to", format(TEST_END),
    "—", n_distinct(returns_test$date), "trading days\n")
cat("LEAKAGE CHECK PASS: No date overlap.\n")
```

```{r feature-engineering}
#| include: false

# ── Rolling helper functions ───────────────────────────────────────────────────

roll_cor <- function(x, y, n) {
  slider::slide2_dbl(x, y,
    .f = ~ cor(.x, .y, use = "complete.obs"),
    .before = n - 1L, .complete = TRUE)
}

roll_beta <- function(x, y, n) {
  slider::slide2_dbl(x, y,
    .f = ~ {
      df <- data.frame(x = .x, y = .y)
      df <- df[complete.cases(df), ]
      if (nrow(df) < 10L) return(NA_real_)
      coef(lm(x ~ y, data = df))[2L]
    },
    .before = n - 1L, .complete = TRUE)
}

roll_cum_ret <- function(x, n) {
  slider::slide_dbl(x,
    .f        = ~ prod(1 + .x, na.rm = FALSE) - 1,
    .before   = n - 1L,
    .complete = TRUE)
}

# ── Compute features for all sector ETFs ──────────────────────────────────────
features_daily <- returns_daily |>
  filter(symbol %in% SECTOR_TICKERS) |>
  left_join(spy_rets_wide, by = "date") |>
  group_by(symbol) |>
  arrange(date) |>
  mutate(
    vol_20      = slider::slide_dbl(retClCl, sd, .before = VOL_SHORT - 1L, .complete = TRUE) * sqrt(252),
    vol_60      = slider::slide_dbl(retClCl, sd, .before = VOL_LONG  - 1L, .complete = TRUE) * sqrt(252),
    corr_spy_60 = roll_cor(retClCl, SPY, CORR_WINDOW),
    beta_spy_60 = roll_beta(retClCl, SPY, CORR_WINDOW),
    cum_ret_60  = roll_cum_ret(retClCl, RET_WINDOW),
    spy_cum_60  = roll_cum_ret(SPY, RET_WINDOW),
    rel_ret_60  = cum_ret_60 - spy_cum_60,
    # Also compute Sharpe (for clustering summary, matching class notes style)
    sharpe_60   = slider::slide_dbl(retClCl,
                    ~ if (sd(.x) == 0) NA_real_ else mean(.x) / sd(.x) * sqrt(252),
                    .before = VOL_LONG - 1L, .complete = TRUE)
  ) |>
  ungroup() |>
  select(-SPY, -spy_cum_60)

cat("Feature engineering complete.\n")
```

```{r feature-check}
#| include: false

features_daily |>
  filter(date >= TRAIN_START) |>
  group_by(symbol) |>
  summarise(
    pct_vol20    = round(mean(!is.na(vol_20)) * 100, 1),
    pct_corr60   = round(mean(!is.na(corr_spy_60)) * 100, 1),
    pct_cumret60 = round(mean(!is.na(cum_ret_60)) * 100, 1),
    .groups = "drop"
  ) |>
  gt() |>
  tab_header(title = "Feature Completeness in Training Period (%)") |>
  cols_label(symbol = "Ticker", pct_vol20 = "vol_20 %",
             pct_corr60 = "corr_spy_60 %", pct_cumret60 = "cum_ret_60 %")
```

```{r feature-dist}
#| include: false

features_daily |>
  filter(date >= TRAIN_START, date <= TRAIN_END) |>
  select(symbol, vol_20, vol_60, corr_spy_60, beta_spy_60, rel_ret_60) |>
  pivot_longer(-symbol, names_to = "feature", values_to = "value") |>
  drop_na() |>
  ggplot(aes(x = value, fill = feature)) +
  geom_histogram(bins = 60, show.legend = FALSE, color = "white", linewidth = 0.1) +
  facet_wrap(~ feature, scales = "free", ncol = 3) +
  scale_fill_viridis_d() +
  labs(title = "Feature Distributions — Training Period (2000–2024)",
       x = "Value", y = "Count") +
  theme_minimal(base_size = 11)
```

```{r month-ends}
#| include: false

# Identify the last SPY trading day in each calendar month.
all_trading_days <- sort(unique(returns_daily$date))

month_ends_all <- tibble(date = all_trading_days) |>
  mutate(ym = floor_date(date, "month")) |>
  group_by(ym) |>
  filter(date == max(date)) |>
  ungroup() |>
  pull(date)

train_month_ends <- month_ends_all[month_ends_all >= TRAIN_START & month_ends_all <= TRAIN_END]
test_month_ends  <- month_ends_all[month_ends_all >= TEST_START  & month_ends_all <= TEST_END]
all_month_ends   <- c(train_month_ends, test_month_ends)

cat("Training rebalance dates:", length(train_month_ends), "\n")
cat("Testing  rebalance dates:", length(test_month_ends), "\n")
```

```{r clustering-functions}
#| include: false

# ── Standardize features using recipes pipeline (tidymodels preprocessing) ────
standardize_features <- function(df) {
  rec <- recipe(~ ., data = df) |>
    step_normalize(all_numeric_predictors()) |>
    prep(training = df, verbose = FALSE)
  bake(rec, new_data = df)
}

# ── Hierarchical clustering (Ward.D2) ─────────────────────────────────────────
cluster_hierarchical <- function(feature_mat, k = N_CLUSTERS, method = LINK_METHOD) {
  scaled     <- standardize_features(feature_mat)
  # recipes::bake() strips rownames → dist() gets no labels → cutree() returns
  # an unnamed vector → names(assignments) == NULL → downstream subsetting breaks.
  # Restore rownames before computing distances so cutree() is ETF-symbol named.
  scaled_mat <- as.matrix(scaled)
  rownames(scaled_mat) <- rownames(feature_mat)
  dist_mat <- dist(scaled_mat, method = "euclidean")
  hc       <- hclust(dist_mat, method = method)
  list(assignments = cutree(hc, k = k), hclust_obj = hc, dist_mat = dist_mat)
}

# ── K-Means clustering ────────────────────────────────────────────────────────
cluster_kmeans <- function(feature_mat, k = N_CLUSTERS, seed = 42L) {
  scaled     <- standardize_features(feature_mat)
  # recipes::bake() drops rownames — restore them so kmeans() assignments
  # are named by ETF symbol and can be indexed by name downstream.
  scaled_mat <- as.matrix(scaled)
  rownames(scaled_mat) <- rownames(feature_mat)
  set.seed(seed)
  km <- stats::kmeans(scaled_mat, centers = k, nstart = 25L)
  list(assignments = km$cluster, kmeans_obj = km, scaled = scaled_mat)
}

# ── Apply clustering at a single rebalance date ────────────────────────────────
cluster_at_date <- function(d, features_df, k = N_CLUSTERS) {
  snap <- features_df |>
    filter(date == d) |>
    select(symbol, all_of(CLUSTER_FEATURES)) |>
    drop_na()

  if (nrow(snap) < k) {
    message("Date ", d, ": only ", nrow(snap), " ETFs — skipping.")
    return(NULL)
  }

  feat_mat <- as.data.frame(select(snap, -symbol))
  rownames(feat_mat) <- snap$symbol

  hc_res <- cluster_hierarchical(feat_mat, k = k)
  km_res <- cluster_kmeans(feat_mat, k = k)

  tibble(
    date           = d,
    symbol         = names(hc_res$assignments),
    cluster        = as.integer(hc_res$assignments),
    cluster_kmeans = as.integer(km_res$assignments[names(hc_res$assignments)])
  )
}
```

```{r run-clustering}
#| include: false

cat("Running monthly clustering across", length(all_month_ends), "rebalance dates...\n")

cluster_history <- purrr::map_dfr(
  all_month_ends,
  ~ cluster_at_date(.x, features_daily, k = N_CLUSTERS)
)

# Coverage check
cluster_coverage <- cluster_history |>
  group_by(date) |>
  summarise(n_etfs = n(), n_clusters = n_distinct(cluster), .groups = "drop")

cat("Clustering complete.\n")
cat("Total assignment rows       :", nrow(cluster_history), "\n")
cat("Avg ETFs per rebalance date :", round(mean(cluster_coverage$n_etfs), 1), "\n")
```

```{r residual-construction}
#| include: false

# ── Step 1: Carry cluster assignments forward to all daily dates ───────────────
symbol_dates <- features_daily |>
  filter(date >= TRAIN_START) |>
  distinct(date, symbol)

cluster_daily <- symbol_dates |>
  left_join(cluster_history |> select(date, symbol, cluster), by = c("date", "symbol")) |>
  arrange(symbol, date) |>
  group_by(symbol) |>
  fill(cluster, .direction = "down") |>
  ungroup()

# ── Step 2: Join cluster assignments to feature data ──────────────────────────
residuals_daily <- features_daily |>
  filter(date >= TRAIN_START) |>
  select(symbol, date, cum_ret_60) |>
  left_join(cluster_daily, by = c("symbol", "date")) |>
  filter(!is.na(cluster), !is.na(cum_ret_60))

# ── Step 3: Leave-one-out cluster mean ───────────────────────────────────────
residuals_daily <- residuals_daily |>
  group_by(date, cluster) |>
  mutate(
    cluster_sum = sum(cum_ret_60, na.rm = TRUE),
    cluster_n   = sum(!is.na(cum_ret_60)),
    peer_mean   = (cluster_sum - cum_ret_60) / (cluster_n - 1L)
  ) |>
  ungroup() |>
  mutate(residual = cum_ret_60 - peer_mean)

solo_count <- residuals_daily |> filter(cluster_n == 1L) |> nrow()
cat("Solo cluster rows (peer_mean = NA):", solo_count, "\n")
cat("Valid residual rows:", sum(!is.na(residuals_daily$residual)), "\n")
```

```{r kalman-function}
#| include: false

#' Local-level Kalman filter (sequential, causal)
#'
#' @param y  Numeric vector: observed residual series. NAs propagated.
#' @param Q  Process noise variance (state drift speed)
#' @param R  Observation noise variance (smoothing strength)
#' @return List: state, pred_state, pred_var, innovation, z_score
kalman_local_level <- function(y, Q = KF_Q, R = KF_R) {
  n        <- length(y)
  state    <- numeric(n)
  P_filt   <- numeric(n)
  pred_st  <- numeric(n)
  pred_var <- numeric(n)
  innov    <- numeric(n)
  z_score  <- numeric(n)

  first_obs <- which(!is.na(y))[1L]
  if (is.na(first_obs)) {
    return(list(state = rep(NA_real_, n), pred_state = rep(NA_real_, n),
                pred_var = rep(NA_real_, n), innovation = rep(NA_real_, n),
                z_score = rep(NA_real_, n)))
  }

  if (first_obs > 1L) {
    state[seq_len(first_obs - 1L)]    <- NA_real_
    P_filt[seq_len(first_obs - 1L)]   <- NA_real_
    pred_st[seq_len(first_obs - 1L)]  <- NA_real_
    pred_var[seq_len(first_obs - 1L)] <- NA_real_
    innov[seq_len(first_obs - 1L)]    <- NA_real_
    z_score[seq_len(first_obs - 1L)]  <- NA_real_
  }

  state[first_obs]    <- y[first_obs]
  P_filt[first_obs]   <- R
  pred_st[first_obs]  <- y[first_obs]
  pred_var[first_obs] <- R + Q
  innov[first_obs]    <- 0
  z_score[first_obs]  <- 0

  for (t in seq(first_obs + 1L, n)) {
    a_pred <- state[t - 1L]
    P_pred <- P_filt[t - 1L] + Q

    if (is.na(y[t])) {
      state[t]    <- a_pred
      P_filt[t]   <- P_pred
      pred_st[t]  <- a_pred
      pred_var[t] <- P_pred + R
      innov[t]    <- NA_real_
      z_score[t]  <- NA_real_
    } else {
      F_t         <- P_pred + R
      v_t         <- y[t] - a_pred
      K_t         <- P_pred / F_t

      state[t]    <- a_pred + K_t * v_t
      P_filt[t]   <- (1 - K_t) * P_pred
      pred_st[t]  <- a_pred
      pred_var[t] <- F_t
      innov[t]    <- v_t
      z_score[t]  <- v_t / sqrt(F_t)
    }
  }

  list(state = state, pred_state = pred_st, pred_var = pred_var,
       innovation = innov, z_score = z_score)
}
```

```{r apply-kalman}
#| include: false

# Filter runs sequentially over full sample (train + test) — strictly causal.
cat("Applying Kalman filter to", n_distinct(residuals_daily$symbol), "ETF residuals...\n")

kalman_results <- residuals_daily |>
  filter(!is.na(residual)) |>
  arrange(symbol, date) |>
  group_by(symbol) |>
  group_modify(~ {
    kf <- kalman_local_level(.x$residual, Q = KF_Q, R = KF_R)
    mutate(.x, kf_state = kf$state, kf_pred_st = kf$pred_state,
           kf_pred_var = kf$pred_var, kf_innov = kf$innovation, kf_z = kf$z_score)
  }) |>
  ungroup()

cat("Kalman filter applied. Valid z-scores:", sum(!is.na(kalman_results$kf_z)), "\n")
```

```{r monthly-returns}
#| include: false

monthly_prices <- prices |>
  filter(date %in% month_ends_all) |>
  arrange(symbol, date)

monthly_ret_tbl <- monthly_prices |>
  group_by(symbol) |>
  arrange(date) |>
  mutate(monthly_ret = adjusted / dplyr::lag(adjusted) - 1) |>
  ungroup() |>
  filter(!is.na(monthly_ret))

cat("Monthly returns computed for", n_distinct(monthly_ret_tbl$symbol), "symbols.\n")
```

```{r trade-stats-function}
#| include: false

# Custom tradeStats function matching RTL::tradeStats() output format
trade_stats <- function(ret_vec, rf = 0) {
  r       <- ret_vec[!is.na(ret_vec)]
  n       <- length(r)
  if (n < 3L) return(tibble())

  cum_ret  <- prod(1 + r) - 1
  ret_ann  <- prod(1 + r)^(12 / n) - 1
  sd_ann   <- sd(r) * sqrt(12)
  sharpe   <- (ret_ann - rf) / sd_ann

  # Omega (MAR = 0, matching class notes)
  MAR       <- 0
  gains     <- r[r > MAR]
  losses    <- r[r <= MAR]
  omega     <- if (length(losses) > 0 && sum(abs(losses)) > 0)
                 sum(gains - MAR) / sum(MAR - losses) else Inf

  pct_win   <- mean(r > 0)
  pct_in_mkt <- mean(r != 0)

  # Max drawdown depth and length
  cum_growth <- cumprod(1 + r)
  running_max <- cummax(cum_growth)
  dd          <- (cum_growth - running_max) / running_max
  dd_max      <- min(dd)

  # Drawdown length: max consecutive periods below peak
  below_peak <- dd < 0
  dd_lengths <- rle(below_peak)
  dd_length  <- if (any(dd_lengths$values)) max(dd_lengths$lengths[dd_lengths$values]) else 0L

  tibble(
    CumReturn  = cum_ret,
    Ret.Ann    = ret_ann,
    SD.Ann     = sd_ann,
    Sharpe     = sharpe,
    Omega      = omega,
    `%.Win`    = pct_win,
    `%.InMrkt` = pct_in_mkt,
    DD.Length  = dd_length,
    DD.Max     = dd_max,
    N.Trades   = sum(r != 0),   # Months with an active position
    N.Months   = n
  )
}
```

# Executive Summary

This project implements a cluster-relative mean reversion strategy on the eleven S&P 500 Select Sector SPDR ETFs. Clustering is the primary signal engine — it groups sectors by behavioral similarity and defines what a fair price comparison means. Temporary deviations of an ETF from its cluster peer group are refined by a local-level Kalman filter, which outputs a standardized z-score used to trigger long and short positions. The model trains on 24 years of daily data (2000–2024) and is evaluated on a 15-month out-of-sample test period (January 2025 – March 2026). All data is sourced via the Yahoo Finance API through R.

# The Problem: Why Cluster Before Comparing?

Sector ETFs are not interchangeable. Energy, Technology, and Utilities respond to different economic forces, carry different volatility profiles, and behave differently across market cycles. A mean reversion strategy that compares any ETF against any other will generate false signals constantly — not because mean reversion is wrong, but because the comparison group is wrong.

The correct question is not whether XLE is above the market average. The correct question is whether XLE has deviated from sectors that actually behave like it. That is a peer-relative valuation problem, and clustering is the natural tool for solving it.

By grouping sectors into behavioral clusters first — and then measuring each ETF's deviation relative to its own cluster — the strategy ensures that mean reversion signals reflect genuine short-term overshoot within a meaningful comparison group. The mental model that follows formalizes this logic and drives every implementation decision.

# Mental Model and Strategy Logic

The diagram below captures the complete decision logic, from problem framing through signal generation to out-of-sample validation. Every component in the R implementation traces back to a specific node in this model. Clustering defines comparability. The Kalman filter refines it. The z-score triggers it. The decision rules execute it.



:::{.mermaid-container}
```{mermaid}
%%| fig-width: 10
%%| fig-height: 16
%%{init: {"theme": "base", "themeVariables": {"fontSize": "14px", "fontFamily": "Inter, sans-serif"}}}%%
flowchart TD
    classDef probClass   fill:#DBEAFE,stroke:#1D4ED8,stroke-width:2px,color:#1e3a5f
    classDef mentClass   fill:#EDE9FE,stroke:#6D28D9,stroke-width:2px,color:#3b0764
    classDef mktClass    fill:#DCFCE7,stroke:#15803D,stroke-width:2px,color:#14532d
    classDef inpClass    fill:#FEF3C7,stroke:#D97706,stroke-width:2px,color:#78350f
    classDef clsClass    fill:#FEE2E2,stroke:#DC2626,stroke-width:3px,color:#7f1d1d,font-weight:bold
    classDef resClass    fill:#CCFBF1,stroke:#0D9488,stroke-width:2px,color:#134e4a
    classDef kfClass     fill:#E0E7FF,stroke:#4338CA,stroke-width:2px,color:#1e1b4b
    classDef decClass    fill:#FEF9C3,stroke:#CA8A04,stroke-width:2px,color:#713f12
    classDef portClass   fill:#D1FAE5,stroke:#059669,stroke-width:2px,color:#064e3b
    classDef valClass    fill:#F1F5F9,stroke:#475569,stroke-width:2px,color:#1e293b

    subgraph P["Problem Framing"]
        P1["Can temporary divergences among sector ETFs<br>be traded when sectors are first grouped<br>into behaviorally similar clusters?"]:::probClass
        P2["Constraints: Daily data · Monthly rebalance<br>SPY + Select Sector ETFs · R / Quarto<br>Clustering primary · Kalman required<br>Code must match mental model"]:::probClass
    end

    subgraph MM["Mental Model Toolbox"]
        M1["Systems Thinking<br>ETFs are interconnected,<br>not isolated assets"]:::mentClass
        M2["Inversion<br>What causes false signals<br>or model failure?"]:::mentClass
        M3["Margin of Safety<br>Only trade sufficiently<br>large deviations"]:::mentClass
        M4["Second-Order Thinking<br>Distinguish temporary noise<br>from structural change"]:::mentClass
        M5["Risk–Reward Framework<br>Signal must justify risk,<br>drawdown, and turnover"]:::mentClass
        M6["Opportunity Cost<br>Capital to strongest<br>signals only"]:::mentClass
    end

    subgraph MV["Market View / Economic Logic"]
        MV1["Sectors form behavioral clusters<br>based on rolling return patterns"]:::mktClass
        MV2["Within a cluster, one ETF can temporarily<br>overshoot or undershoot peers"]:::mktClass
        MV3["Fair value is cluster-relative,<br>not global — apples to apples only"]:::mktClass
    end

    subgraph DI["Daily Inputs"]
        D1["Adjusted daily prices & returns"]:::inpClass
        D2["Rolling 20-day & 60-day volatility"]:::inpClass
        D3["Rolling 60-day correlation to SPY"]:::inpClass
        D4["Rolling 60-day relative return vs SPY"]:::inpClass
        D5["Rolling 60-day beta to SPY"]:::inpClass
    end

    subgraph CL["PRIMARY SIGNAL ENGINE: Clustering"]
        C1["Elbow method → choose k clusters"]:::clsClass
        C2["K-Means + Hierarchical Ward.D2<br>Agreement = robust structure"]:::clsClass
        C3["Silhouette validation<br>Silhouette > 0.5 = well-clustered"]:::clsClass
        C4["Fair value is cluster-relative, not global"]:::clsClass
    end

    subgraph RE["Residual Construction"]
        R1["60-day compounded return for each ETF"]:::resClass
        R2["Cluster benchmark =<br>leave-one-out peer mean return"]:::resClass
        R3["Residual = ETF cum return − cluster benchmark"]:::resClass
    end

    subgraph KF["Kalman Filter — Smoothed State Estimate"]
        K1["Local-level state-space model<br>applied to each ETF's residual series"]:::kfClass
        K2["Estimate latent fair-value equilibrium<br>smooth noise, improve reliability"]:::kfClass
        K3["Standardized innovation = z-score signal<br>Kalman refines, does not replace clustering"]:::kfClass
    end

    subgraph DR["Decision Rules"]
        DR1{"|z-score| ≥ threshold τ?"}:::decClass
        DR2["z < −τ → LONG<br>ETF below Kalman fair value<br>Expect reversion upward"]:::decClass
        DR3["z > +τ → SHORT<br>ETF above Kalman fair value<br>Expect reversion downward"]:::decClass
        DR4["|z| < τ → NEUTRAL<br>Deviation too small<br>Margin of safety not met"]:::decClass
        DR5["Rank signals by |z|<br>Hold only strongest N per side"]:::decClass
    end

    subgraph PC["Portfolio Construction"]
        PC1["Monthly rebalance at month-end close"]:::portClass
        PC2["Equal-weight long & short positions"]:::portClass
        PC3["Signal threshold = margin of safety<br>Max positions = opportunity cost filter"]:::portClass
    end

    subgraph VA["Validation & Iteration"]
        V1["Training 2000–2024: parameter selection"]:::valClass
        V2["Testing 2025–2026: out-of-sample evaluation"]:::valClass
        V3["tradeStats: Sharpe, Omega, %.Win,<br>%.InMrkt, DD.Max, DD.Length"]:::valClass
        V4["Bias checks: look-ahead, overfitting,<br>storytelling, structural breaks"]:::valClass
    end

    style P  fill:#EFF6FF,stroke:#1D4ED8,stroke-width:2px
    style MM fill:#F5F3FF,stroke:#6D28D9,stroke-width:2px
    style MV fill:#F0FDF4,stroke:#15803D,stroke-width:2px
    style DI fill:#FFFBEB,stroke:#D97706,stroke-width:2px
    style CL fill:#FFF1F2,stroke:#DC2626,stroke-width:3px
    style RE fill:#F0FDFA,stroke:#0D9488,stroke-width:2px
    style KF fill:#EEF2FF,stroke:#4338CA,stroke-width:2px
    style DR fill:#FEFCE8,stroke:#CA8A04,stroke-width:2px
    style PC fill:#ECFDF5,stroke:#059669,stroke-width:2px
    style VA fill:#F8FAFC,stroke:#475569,stroke-width:2px

    P --> MM --> MV --> DI --> CL --> RE --> KF --> DR
    DR --> DR1
    DR1 --> DR2
    DR1 --> DR3
    DR1 --> DR4
    DR2 & DR3 --> DR5 --> PC --> VA
    VA -->|"Reflect & refine parameters"| CL
```

:::

```{=html}
<script>
(function () {
  var scale = 1;
  function getSvg() {
    return document.querySelector('.mermaid-container svg');
  }
  function applyScale() {
    var svg = getSvg();
    if (!svg) return;
    svg.style.transform       = 'scale(' + scale + ')';
    svg.style.transformOrigin = 'top left';
    svg.style.display         = 'block';
    var lbl = document.getElementById('mzoom-label');
    if (lbl) lbl.textContent = Math.round(scale * 100) + '%';
  }
  window.mzoom = function (f) {
    scale = Math.max(0.2, Math.min(scale * f, 5));
    applyScale();
  };
  window.mreset = function () {
    scale = 1;
    applyScale();
  };
  /* Poll until mermaid has rendered the SVG (client-side rendering) */
  var tries = 0;
  var timer = setInterval(function () {
    if (getSvg() || ++tries > 60) { clearInterval(timer); applyScale(); }
  }, 100);
})();
</script>
```

# Data, Universe, and Constraints

The universe consists of SPY as the benchmark and eleven Select Sector SPDR ETFs covering the full S&P 500 sector breakdown. Two ETFs — XLC (Communication Services, live June 2018) and XLRE (Real Estate, live October 2015) — have shorter histories and are excluded from analysis prior to their first trading day. All prices are downloaded from Yahoo Finance via `tidyquant::tq_get()` using adjusted close prices that account for dividends and splits. The training window spans June 2000 to December 2024 (approximately 294 monthly rebalance dates). The out-of-sample test window covers January 2025 to March 2026 (15 months). No testing-period data is used during model design or parameter selection.

```{r data-checks}
# ── Check 1: All tickers present ──────────────────────────────────────────────
missing_tickers <- setdiff(TICKERS, unique(raw_prices$symbol))
if (length(missing_tickers) > 0) {
  stop("Missing tickers: ", paste(missing_tickers, collapse = ", "))
} else {
  cat("CHECK 1 PASS: All", length(TICKERS), "tickers downloaded.\n")
}

# ── Check 2: Date coverage and NA rates ───────────────────────────────────────
coverage_tbl <- raw_prices |>
  group_by(symbol) |>
  summarise(
    first_date = min(date),
    last_date  = max(date),
    n_obs      = n(),
    pct_na_adj = round(mean(is.na(adjusted)) * 100, 2),
    .groups    = "drop"
  ) |>
  arrange(first_date)

coverage_tbl |>
  gt() |>
  tab_header(title = "Data Coverage by Ticker") |>
  fmt_date(columns = c(first_date, last_date), date_style = 2) |>
  fmt_number(columns = pct_na_adj, decimals = 2) |>
  cols_label(
    symbol     = "Ticker",
    first_date = "First Date",
    last_date  = "Last Date",
    n_obs      = "Trading Days",
    pct_na_adj = "% Missing (Adj)"
  )

# ── Check 3: No NA in adjusted for standard tickers in training period ─────────
standard_tickers <- setdiff(TICKERS, c("XLC", "XLRE"))
na_standard <- raw_prices |>
  filter(symbol %in% standard_tickers, date >= TRAIN_START, is.na(adjusted)) |>
  nrow()

if (na_standard > 0) {
  warning("NA values found in adjusted prices for standard tickers in training period!")
} else {
  cat("CHECK 3 PASS: No NA in adjusted prices for standard tickers (training period).\n")
}
```

# Strategy Design and Decision Rules

**Step 1 — Feature Construction.** For each sector ETF, compute five rolling behavioral features at each trading day: 20-day annualized volatility, 60-day annualized volatility, 60-day SPY correlation, 60-day SPY beta, and 60-day relative return versus SPY. Features are standardized using a `recipes` normalization pipeline before any clustering step.

**Step 2 — Cluster Assignment.** At each month-end, run both Hierarchical clustering (Ward.D2 linkage) and K-Means clustering on the standardized feature snapshot. Agreement between the two methods confirms robust cluster structure. The number of clusters is selected using the WSS elbow method and validated by silhouette analysis.

**Step 3 — Cluster-Relative Residual.** Compute the 60-day compounded return for each ETF and the leave-one-out cluster peer mean (average of all other ETFs in the same cluster, excluding self to avoid inflation). The residual is: ETF 60-day return − leave-one-out peer mean.

**Step 4 — Kalman Filter.** Apply a local-level state-space Kalman filter (random walk + noise model) to each ETF's daily residual series. The filter estimates a latent fair-value equilibrium. The standardized innovation — the z-score — measures how surprising the current deviation is relative to the filter's own uncertainty.

**Step 5 — Decision Rules.**

| Condition | Direction | Interpretation |
|---|---|---|
| z < −τ | **LONG** | ETF below cluster fair value; mean reversion expected upward |
| z > +τ | **SHORT** | ETF above cluster fair value; mean reversion expected downward |
| \|z\| < τ | **NEUTRAL** | Deviation too small; margin of safety not met |

Rank all long candidates by z (most negative first) and shorts by z (most positive first). Hold the strongest MAX_LONGS = 3 longs and MAX_SHORTS = 3 shorts, equal-weighted within each book. Rebalance at every month-end close. The following month's close-to-close return is realized.

```{r elbow-method}
# Compute within-cluster sum of squares for k = 1 to 6 on a representative snapshot
# Use the most recent training month-end for illustration
last_train_date <- max(train_month_ends)

snap_elbow <- features_daily |>
  filter(date == last_train_date) |>
  select(symbol, all_of(CLUSTER_FEATURES)) |>
  drop_na()

feat_elbow <- as.data.frame(select(snap_elbow, -symbol))
rownames(feat_elbow) <- snap_elbow$symbol

# Standardize (matching recipes::recipe() approach)
scaled_elbow <- scale(feat_elbow)

max_k <- min(6L, nrow(scaled_elbow) - 1L)

wss <- purrr::map_dbl(1:max_k, function(k) {
  stats::kmeans(scaled_elbow, centers = k, nstart = 25L)$tot.withinss
})

tibble(k = 1:max_k, wss = wss) |>
  ggplot(aes(x = k, y = wss)) +
  geom_line(color = "#1565C0", linewidth = 1) +
  geom_point(size = 3, color = "#1565C0") +
  geom_vline(xintercept = N_CLUSTERS, linetype = "dashed", color = "#E53935") +
  scale_x_continuous(breaks = 1:max_k) +
  labs(
    title    = "Elbow Method — Optimal Number of Clusters",
    subtitle = paste0("Red dashed line = chosen k = ", N_CLUSTERS,
                      " | Look for the elbow where improvement slows"),
    x = "Number of Clusters (k)", y = "Within-Cluster Sum of Squares (WSS)"
  ) +
  theme_minimal(base_size = 11)
```

```{r silhouette-analysis}
# Compute silhouette for K-Means and Hierarchical on the most recent training snapshot
snap_sil <- features_daily |>
  filter(date == last_train_date) |>
  select(symbol, all_of(CLUSTER_FEATURES)) |>
  drop_na()

feat_sil <- as.data.frame(select(snap_sil, -symbol))
rownames(feat_sil) <- snap_sil$symbol
scaled_sil  <- as.matrix(standardize_features(feat_sil))
dist_sil    <- dist(scaled_sil)

# Hierarchical silhouette
hc_sil  <- hclust(dist_sil, method = LINK_METHOD)
cuts_hc <- cutree(hc_sil, k = N_CLUSTERS)
sil_hc  <- cluster::silhouette(cuts_hc, dist_sil)

# K-Means silhouette
set.seed(42L)
km_sil  <- stats::kmeans(scaled_sil, centers = N_CLUSTERS, nstart = 25L)
sil_km  <- cluster::silhouette(km_sil$cluster, dist_sil)

p_sil_hc <- factoextra::fviz_silhouette(sil_hc) +
  labs(title = paste("Silhouette — Hierarchical (k =", N_CLUSTERS, ")"),
       subtitle = paste("Avg width:", round(mean(sil_hc[, "sil_width"]), 3),
                        "| >0.5 = well-clustered | <0.3 = weak")) +
  theme_minimal(base_size = 10)

p_sil_km <- factoextra::fviz_silhouette(sil_km) +
  labs(title = paste("Silhouette — K-Means (k =", N_CLUSTERS, ")"),
       subtitle = paste("Avg width:", round(mean(sil_km[, "sil_width"]), 3))) +
  theme_minimal(base_size = 10)

p_sil_hc / p_sil_km

# Agreement check — do both methods agree?
agreement_pct <- mean(cuts_hc == km_sil$cluster) * 100
cat("\nCluster method agreement:", round(agreement_pct, 1), "%\n")
cat("Interpretation: Agreement >= 70% suggests robust structure.\n")
```

```{r method-comparison}
# Compare cluster assignments at the last training date
cluster_history |>
  filter(date == last_train_date) |>
  select(symbol, cluster_hclust = cluster, cluster_kmeans) |>
  mutate(agree = if_else(cluster_hclust == cluster_kmeans, "✓ Agree", "✗ Differ")) |>
  gt() |>
  tab_header(
    title    = "K-Means vs Hierarchical Cluster Assignments",
    subtitle = paste("As of", format(last_train_date))
  ) |>
  cols_label(
    symbol         = "ETF",
    cluster_hclust = "Hierarchical",
    cluster_kmeans = "K-Means",
    agree          = "Agreement"
  ) |>
  tab_style(
    style     = cell_fill(color = "#E8F5E9"),
    locations = cells_body(rows = agree == "✓ Agree")
  ) |>
  tab_style(
    style     = cell_fill(color = "#FFEBEE"),
    locations = cells_body(rows = agree == "✗ Differ")
  )
```

```{r cluster-heatmap}
cluster_history |>
  filter(date >= TRAIN_START, date <= TRAIN_END) |>
  mutate(cluster = factor(cluster)) |>
  ggplot(aes(x = date, y = symbol, fill = cluster)) +
  geom_tile(color = "white", linewidth = 0.15) +
  scale_fill_viridis_d(name = "Cluster") +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  labs(
    title    = "Cluster Assignments Over Time — Training Period",
    subtitle = "Each cell = hierarchical cluster label at that monthly rebalance",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

```{r dendrogram}
snap_last <- features_daily |>
  filter(date == last_train_date) |>
  select(symbol, all_of(CLUSTER_FEATURES)) |>
  drop_na()

feat_last <- as.data.frame(select(snap_last, -symbol))
rownames(feat_last) <- snap_last$symbol
scaled_last <- standardize_features(feat_last)
hc_last     <- hclust(dist(scaled_last), method = LINK_METHOD)

fviz_dend(
  hc_last,
  k          = N_CLUSTERS,
  cex        = 0.85,
  k_colors   = viridis(N_CLUSTERS),
  rect       = TRUE,
  rect_fill  = TRUE,
  rect_lty   = 2,
  horiz      = TRUE,
  main       = paste("Cluster Dendrogram —", format(last_train_date)),
  sub        = paste0("Ward.D2 linkage | k = ", N_CLUSTERS, " clusters"),
  xlab       = "Height", ylab = ""
)
```

```{r cluster-summary-table}
# Compute summary stats per cluster at the most recent training date
snap_summary <- features_daily |>
  filter(date == last_train_date) |>
  select(symbol, vol_20, vol_60, corr_spy_60, beta_spy_60, rel_ret_60, sharpe_60) |>
  drop_na() |>
  left_join(
    cluster_history |> filter(date == last_train_date) |> select(symbol, cluster),
    by = "symbol"
  ) |>
  filter(!is.na(cluster))

snap_summary |>
  group_by(cluster) |>
  summarise(
    n_sectors   = n(),
    avg_vol20   = mean(vol_20),
    avg_corr    = mean(corr_spy_60),
    avg_beta    = mean(beta_spy_60),
    avg_sharpe  = mean(sharpe_60, na.rm = TRUE),
    sectors     = paste(symbol, collapse = ", "),
    .groups     = "drop"
  ) |>
  gt() |>
  tab_header(
    title    = "Cluster Characteristics Summary",
    subtitle = paste("Hierarchical, Ward.D2 | k =", N_CLUSTERS, "| as of", format(last_train_date))
  ) |>
  data_color(
    columns = c(avg_vol20, avg_corr, avg_beta, avg_sharpe),
    palette = c("red", "yellow", "green")
  ) |>
  fmt_percent(columns = avg_vol20, decimals = 1) |>
  fmt_number(columns = c(avg_corr, avg_beta, avg_sharpe), decimals = 2) |>
  fmt_integer(columns = n_sectors) |>
  cols_label(
    cluster    = "Cluster",
    n_sectors  = "# ETFs",
    avg_vol20  = "Ann. Vol (20d)",
    avg_corr   = "Corr to SPY",
    avg_beta   = "Beta to SPY",
    avg_sharpe = "Rolling Sharpe",
    sectors    = "Members"
  )
```

```{r cluster-stability}
# Pre-2020 (excluding COVID) vs Post-2020 feature profiles
pre_2020_date  <- month_ends_all[month_ends_all >= "2019-01-01" &
                                  month_ends_all <= "2019-12-31"] |> max()
post_2020_date <- month_ends_all[month_ends_all >= "2021-01-01" &
                                  month_ends_all <= "2021-12-31"] |> max()

make_regime_snap <- function(d, label) {
  features_daily |>
    filter(date == d) |>
    select(symbol, vol_20, corr_spy_60) |>
    drop_na() |>
    mutate(period = label, date = d)
}

bind_rows(
  make_regime_snap(pre_2020_date,  "Pre-COVID (2019)"),
  make_regime_snap(post_2020_date, "Post-COVID (2021)")
) |>
  ggplot(aes(x = corr_spy_60, y = vol_20, color = period)) +
  geom_point(size = 3.5, alpha = 0.8) +
  geom_text(aes(label = symbol), vjust = -0.8, hjust = 0.5, size = 3) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_color_manual(values = c("Pre-COVID (2019)" = "#1565C0",
                                "Post-COVID (2021)" = "#E65100")) +
  facet_wrap(~ period) +
  labs(
    title    = "Cluster Stability Across Market Regimes",
    subtitle = "Volatility vs SPY correlation — do sector groupings persist?",
    x = "Rolling Correlation to SPY", y = "Ann. Volatility (20d)", color = NULL
  ) +
  theme_minimal(base_size = 11)
```

# Signal Construction: Residuals and Kalman Filtering

The cluster-relative residuals shown below capture each ETF's cumulative deviation from its behavioral peers over a 60-day window. These residuals are noisy — the Kalman filter separates genuine deviations from random fluctuation. The diagnostic charts confirm that the Kalman innovations approximate a standard normal distribution, validating the filter specification. The z-score threshold marks the margin of safety: only deviations exceeding τ standard deviations of innovation trigger a trade.

```{r residual-viz}
residuals_daily |>
  filter(date >= TRAIN_START, date <= TRAIN_END, !is.na(residual)) |>
  ggplot(aes(x = date, y = residual, color = symbol)) +
  geom_line(alpha = 0.7, linewidth = 0.35) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", linewidth = 0.4) +
  facet_wrap(~ symbol, ncol = 3, scales = "free_y") +
  scale_color_viridis_d(guide = "none") +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  labs(
    title    = "Cluster-Relative Residual Series — Training Period",
    subtitle = "Residual = 60-day compounded return minus leave-one-out cluster peer mean",
    x = NULL, y = "Residual"
  ) +
  theme_minimal(base_size = 10) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r kalman-viz}
# Observed vs filtered state for selected ETFs
selected_etfs <- c("XLE", "XLK", "XLF", "XLU")

kalman_results |>
  filter(symbol %in% selected_etfs, date >= TRAIN_START, date <= TRAIN_END) |>
  pivot_longer(c(residual, kf_state, kf_pred_st),
               names_to = "series", values_to = "value") |>
  mutate(series = recode(series, residual = "Observed Residual",
                         kf_state = "Kalman Filtered State",
                         kf_pred_st = "Predicted State")) |>
  ggplot(aes(x = date, y = value, color = series, alpha = series)) +
  geom_line(linewidth = 0.45) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.3) +
  scale_color_manual(values = c("grey55", "#1565C0", "#E53935")) +
  scale_alpha_manual(values = c(0.4, 0.95, 0.7), guide = "none") +
  facet_wrap(~ symbol, ncol = 2, scales = "free_y") +
  labs(title = "Kalman Filter: Observed Residual vs Smoothed Fair-Value State",
       subtitle = "Blue = filtered state | Grey = observed | Red = predicted",
       x = NULL, y = "Residual", color = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

# Innovation distribution — should approximate N(0,1) if well-specified
kalman_results |>
  filter(date >= TRAIN_START, date <= TRAIN_END, !is.na(kf_z)) |>
  ggplot(aes(x = kf_z)) +
  geom_histogram(aes(y = after_stat(density)), bins = 70,
                 fill = "#1565C0", alpha = 0.7, color = "white", linewidth = 0.1) +
  stat_function(fun = dnorm, color = "#E53935", linewidth = 1) +
  geom_vline(xintercept = c(-SIGNAL_THRESHOLD, SIGNAL_THRESHOLD),
             linetype = "dashed", color = "#FF8F00", linewidth = 0.9) +
  labs(title = "Distribution of Kalman Standardised Innovations — Training Period",
       subtitle = paste0("Orange dashed = ±", SIGNAL_THRESHOLD,
                         " threshold | Red = N(0,1)"),
       x = "z-score", y = "Density") +
  theme_minimal(base_size = 11)
```

# Portfolio Construction

At each month-end, signals are ranked by z-score magnitude. Long positions are equal-weighted within the long book; short positions are equal-weighted within the short book. The strategy holds at most 3 long and 3 short positions per rebalance. Months with no qualifying signals produce zero returns — idle capital is preferable to underpowered trades. The position weight heatmap below shows which ETFs were held over time; trade markers on the SPY price chart show the rebalance timing.

```{r signal-generation}
signals_monthly <- kalman_results |>
  filter(date %in% all_month_ends, !is.na(kf_z)) |>
  select(date, symbol, residual, kf_state, kf_z, cluster) |>
  arrange(date, desc(abs(kf_z))) |>
  mutate(signal = case_when(
    kf_z < -SIGNAL_THRESHOLD ~ "LONG",
    kf_z >  SIGNAL_THRESHOLD ~ "SHORT",
    TRUE                     ~ "NEUTRAL"
  ))

# Signal distribution (training)
signals_monthly |>
  filter(date >= TRAIN_START, date <= TRAIN_END) |>
  count(signal) |>
  mutate(pct = n / sum(n) * 100) |>
  gt() |>
  tab_header(title = "Signal Distribution — Training Period") |>
  fmt_number(columns = pct, decimals = 1) |>
  cols_label(signal = "Signal", n = "Count", pct = "% of All")

# Signal activity over time
signals_monthly |>
  filter(date >= TRAIN_START, date <= TRAIN_END) |>
  count(date, signal) |>
  ggplot(aes(x = date, y = n, fill = signal)) +
  geom_col(position = "stack", width = 25) +
  scale_fill_manual(values = c(LONG = "#4CAF50", NEUTRAL = "#9E9E9E", SHORT = "#E53935")) +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  labs(title = "Signal Count per Rebalance Date — Training Period",
       subtitle = paste("Threshold |z| ≥", SIGNAL_THRESHOLD),
       x = NULL, y = "Number of ETFs", fill = "Signal") +
  theme_minimal(base_size = 11)
```

```{r portfolio-construction}
# ── Position builder ──────────────────────────────────────────────────────────
build_positions <- function(signals_df,
                            max_longs  = MAX_LONGS,
                            max_shorts = MAX_SHORTS) {
  longs  <- signals_df |> filter(signal == "LONG")  |> arrange(kf_z)       |> head(max_longs)
  shorts <- signals_df |> filter(signal == "SHORT") |> arrange(desc(kf_z)) |> head(max_shorts)
  n_long <- nrow(longs); n_short <- nrow(shorts)

  # If no signals clear the threshold, return a typed empty tibble.
  # bind_rows(tibble(), tibble()) produces a 0-col tibble; select() then
  # errors with "column doesn't exist". The typed return avoids that.
  if (n_long == 0L && n_short == 0L) {
    return(tibble(symbol = character(),
                  signal = character(),
                  kf_z   = numeric(),
                  weight = numeric()))
  }

  bind_rows(
    if (n_long  > 0L) mutate(longs,  weight =  1 / n_long)  else tibble(),
    if (n_short > 0L) mutate(shorts, weight = -1 / n_short) else tibble()
  ) |>
    # Note: 'date' is omitted — group_modify() strips it from .x and re-attaches it.
    # 'cluster' is selected with any_of() so the function works whether or not
    # the caller includes cluster (main backtest does; sensitivity sweeps don't).
    select(symbol, signal, kf_z, any_of("cluster"), weight)
}

positions_all <- signals_monthly |>
  group_by(date) |>
  group_modify(~ build_positions(.x)) |>
  ungroup()

# Portfolio exposure summary
pos_summary <- positions_all |>
  group_by(date) |>
  summarise(n_long  = sum(signal == "LONG"),
            n_short = sum(signal == "SHORT"),
            gross   = sum(abs(weight)),
            net     = sum(weight), .groups = "drop")

cat("Avg longs per month  :", round(mean(pos_summary$n_long), 1), "\n")
cat("Avg shorts per month :", round(mean(pos_summary$n_short), 1), "\n")
cat("Months with no positions :", sum(pos_summary$gross == 0), "\n")

# Gross/net exposure chart
pos_summary |>
  filter(date >= TRAIN_START) |>
  pivot_longer(c(gross, net), names_to = "type", values_to = "exposure") |>
  ggplot(aes(x = date, y = exposure, color = type)) +
  geom_line(linewidth = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = TEST_START, linetype = "dotted", color = "black") +
  scale_color_manual(values = c(gross = "#1565C0", net = "#E53935")) +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  labs(title = "Portfolio Gross and Net Exposure per Rebalance",
       subtitle = "Dotted = start of testing period",
       x = NULL, y = "Exposure", color = "Type") +
  theme_minimal(base_size = 11)
```

```{r position-chart}
# Build a full grid: every sector ETF × every rebalance date (post TRAIN_START)
all_symbols_pos <- expand_grid(
  symbol = SECTOR_TICKERS,
  date   = all_month_ends[all_month_ends >= TRAIN_START]
)

positions_full_grid <- all_symbols_pos |>
  left_join(
    positions_all |> select(date, symbol, weight, signal),
    by = c("date", "symbol")
  ) |>
  mutate(weight = replace_na(weight, 0))

# Heatmap: green = long, red = short, white = flat
ggplot(positions_full_grid, aes(x = date, y = symbol, fill = weight)) +
  geom_tile(color = "white", linewidth = 0.1) +
  scale_fill_gradient2(
    low      = "#E53935",
    mid      = "white",
    high     = "#4CAF50",
    midpoint = 0,
    name     = "Weight",
    labels   = percent_format(accuracy = 1)
  ) +
  geom_vline(xintercept = as.numeric(TEST_START),
             linetype = "dotted", linewidth = 0.7, color = "black") +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  labs(
    title    = "ETF Position Weights Over Time",
    subtitle = "Green = long | Red = short | White = flat | Dotted line = start of testing period",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

```{r trade-markers}
# SPY monthly price as the backdrop
spy_prices_monthly <- monthly_prices |>
  filter(symbol == "SPY", date >= TRAIN_START) |>
  select(date, adjusted)

# Summarise position direction at each rebalance
position_markers <- positions_all |>
  filter(date >= TRAIN_START) |>
  group_by(date) |>
  summarise(
    n_long  = sum(signal == "LONG"),
    n_short = sum(signal == "SHORT"),
    .groups = "drop"
  ) |>
  mutate(direction = case_when(
    n_long > 0 & n_short > 0  ~ "Long & Short",
    n_long > 0                 ~ "Long Only",
    n_short > 0                ~ "Short Only",
    TRUE                       ~ "Flat"
  )) |>
  filter(direction != "Flat") |>
  left_join(spy_prices_monthly, by = "date")

ggplot(spy_prices_monthly, aes(x = date, y = adjusted)) +
  geom_line(color = "grey55", linewidth = 0.5) +
  # Long entries (upward triangles)
  geom_point(
    data  = position_markers |> filter(grepl("Long", direction)),
    aes(color = "Long Entry"),
    shape = 24, size = 2.8, fill = "#4CAF50", stroke = 0.4
  ) +
  # Short entries (downward triangles)
  geom_point(
    data  = position_markers |> filter(grepl("Short", direction)),
    aes(color = "Short Entry"),
    shape = 25, size = 2.8, fill = "#E53935", stroke = 0.4
  ) +
  geom_vline(xintercept = TEST_START, linetype = "dotted",
             color = "black", linewidth = 0.7) +
  annotate("text", x = TEST_START + 30, y = min(spy_prices_monthly$adjusted, na.rm = TRUE),
           label = "← Train | Test →", size = 3, hjust = 0, color = "grey40") +
  scale_color_manual(
    values = c("Long Entry" = "#4CAF50", "Short Entry" = "#E53935"),
    guide  = guide_legend(
      override.aes = list(shape = c(24L, 25L), fill = c("#4CAF50", "#E53935"))
    )
  ) +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  scale_y_continuous(labels = dollar_format()) +
  labs(
    title    = "Trade Markers on SPY Price Chart",
    subtitle = "▲ = rebalance with ≥1 long | ▼ = rebalance with ≥1 short | Monthly frequency",
    x = NULL, y = "SPY Adjusted Price ($)", color = "Trade Direction"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

```{r backtest}
#| include: false

# Signal at month-end d → position held in following month d+1.
# Return(d+1) = strategy weighted ETF returns from close d to close d+1.
next_month_map <- tibble(
  signal_date = all_month_ends[-length(all_month_ends)],
  return_date = all_month_ends[-1]
)

strategy_returns_raw <- positions_all |>
  rename(signal_date = date) |>
  left_join(next_month_map, by = "signal_date") |>
  filter(!is.na(return_date)) |>
  left_join(monthly_ret_tbl |> select(symbol, return_date = date, monthly_ret),
            by = c("symbol", "return_date")) |>
  filter(!is.na(monthly_ret)) |>
  group_by(return_date) |>
  summarise(strategy_ret = sum(weight * monthly_ret, na.rm = TRUE),
            n_positions  = n(), .groups = "drop")

all_return_months <- tibble(return_date = all_month_ends[-1])
strategy_returns <- all_return_months |>
  left_join(strategy_returns_raw, by = "return_date") |>
  mutate(strategy_ret = replace_na(strategy_ret, 0),
         n_positions  = replace_na(n_positions, 0)) |>
  filter(return_date >= TRAIN_START)

spy_monthly <- monthly_ret_tbl |>
  filter(symbol == "SPY") |>
  select(return_date = date, spy_ret = monthly_ret)

backtest_df <- strategy_returns |>
  left_join(spy_monthly, by = "return_date") |>
  filter(!is.na(spy_ret)) |>
  mutate(period = if_else(return_date <= TRAIN_END, "Training", "Testing"))

cat("Training months:", sum(backtest_df$period == "Training"), "\n")
cat("Testing  months:", sum(backtest_df$period == "Testing"), "\n")
```

# Backtest Results

The backtest runs monthly close-to-close from June 2000 through March 2026 with a strict temporal split. No future information enters the signal generation or Kalman initialization. The cumulative equity chart shows both training and testing period performance against SPY. The training/testing facet chart rebases each period to zero for direct comparison. The trade statistics table summarizes the key performance metrics across both periods with SPY as the benchmark.

```{r strategy-chart}
# Net exposure at each return month (sum of weights: +1 = fully long, -1 = fully short)
net_pos_monthly <- positions_all |>
  group_by(date) |>
  summarise(net_pos = sum(weight), .groups = "drop") |>
  rename(signal_date = date) |>
  left_join(next_month_map, by = "signal_date") |>
  filter(!is.na(return_date)) |>
  select(return_date, net_pos)

# Build monthly xts including net position exposure
backtest_xts <- backtest_df |>
  select(return_date, strategy_ret, spy_ret) |>
  left_join(net_pos_monthly, by = "return_date") |>
  mutate(
    cum_strat = cumprod(1 + strategy_ret),
    cum_spy   = cumprod(1 + spy_ret),
    net_pos   = replace_na(net_pos, 0)
  ) |>
  timetk::tk_xts(date_var = return_date)

# 4-panel course-style xts chart matching §8.4.4 / §8.10 format:
#   Panel 1: CumEQ (Strategy vs SPY overlay)
#   Panel 2: Positions — net directional exposure (sum of weights)
#   Panel 3: Monthly Return bars
plot(backtest_xts$cum_strat,
     main  = "Strategy Results",
     col   = "#1565C0",
     ylab  = "Growth of $1")
xts::addSeries(backtest_xts$cum_spy,
     main = "SPY", on = 1,
     type = "l", col = "#E65100", lty = 2, lwd = 1.5, pch = 0)
xts::addSeries(backtest_xts$net_pos,
     main = "Positions",
     on   = NA,
     type = "h",
     col  = "blue",
     lty  = 1,
     lwd  = 1,
     pch  = 0)
xts::addSeries(backtest_xts$strategy_ret,
     main = "Monthly Return",
     on   = NA,
     type = "h",
     col  = "#1565C0",
     lty  = 1,
     lwd  = 1,
     pch  = 0)
```

```{r cum-returns}
backtest_df |>
  select(return_date, strategy_ret, spy_ret) |>
  pivot_longer(-return_date, names_to = "series", values_to = "ret") |>
  group_by(series) |>
  mutate(cum_ret = cumprod(1 + ret) - 1) |>
  ungroup() |>
  mutate(series = recode(series,
    strategy_ret = "Strategy — Cluster Mean Reversion",
    spy_ret      = "SPY Benchmark")) |>
  ggplot(aes(x = return_date, y = cum_ret, color = series, linetype = series)) +
  geom_line(linewidth = 0.9) +
  geom_vline(xintercept = TEST_START, linetype = "dotted",
             color = "black", linewidth = 0.7) +
  annotate("text", x = TEST_START + 20, y = -0.05,
           label = "← Training  |  Testing →", size = 3.2, hjust = 0, color = "grey30") +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_color_manual(values = c("#1565C0", "#E65100")) +
  scale_linetype_manual(values = c("solid", "dashed")) +
  labs(title = "Cumulative Returns: Strategy vs SPY Benchmark",
       x = NULL, y = "Cumulative Return", color = NULL, linetype = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

```{r train-test-comparison}
backtest_df |>
  select(return_date, period, strategy_ret, spy_ret) |>
  group_by(period) |>
  mutate(cum_strat = cumprod(1 + strategy_ret) - 1,
         cum_spy   = cumprod(1 + spy_ret) - 1) |>
  ungroup() |>
  pivot_longer(c(cum_strat, cum_spy), names_to = "series", values_to = "cum_ret") |>
  mutate(series = recode(series, cum_strat = "Strategy", cum_spy = "SPY")) |>
  ggplot(aes(x = return_date, y = cum_ret, color = series)) +
  geom_line(linewidth = 0.85) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~ period, scales = "free") +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_color_manual(values = c(Strategy = "#1565C0", SPY = "#E65100")) +
  labs(title = "Cumulative Returns by Period (Rebased to Zero)",
       x = NULL, y = "Cumulative Return", color = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

```{r trade-stats-output}
# Compute for all period / series combinations
stats_list <- list(
  "Strategy — Training" = filter(backtest_df, period == "Training")$strategy_ret,
  "Strategy — Testing"  = filter(backtest_df, period == "Testing")$strategy_ret,
  "Strategy — Full"     = backtest_df$strategy_ret,
  "SPY — Training"      = filter(backtest_df, period == "Training")$spy_ret,
  "SPY — Testing"       = filter(backtest_df, period == "Testing")$spy_ret,
  "SPY — Full"          = backtest_df$spy_ret
)

stats_tbl <- purrr::imap_dfr(stats_list, ~ mutate(trade_stats(.x), Period = .y, .before = 1))

stats_tbl |>
  gt() |>
  tab_header(
    title    = "Performance Summary — tradeStats",
    subtitle = "Monthly returns | annualised metrics | matching class notes format"
  ) |>
  fmt_percent(columns = c(CumReturn, Ret.Ann, SD.Ann, `%.Win`, `%.InMrkt`, DD.Max),
              decimals = 1) |>
  fmt_number(columns  = c(Sharpe, Omega), decimals = 2) |>
  fmt_integer(columns = c(DD.Length, N.Trades, N.Months)) |>
  tab_style(
    style     = cell_fill(color = "#E3F2FD"),
    locations = cells_body(rows = grepl("Strategy", Period))
  ) |>
  cols_label(
    Period     = "Series",
    CumReturn  = "Cum Return",
    Ret.Ann    = "Ann. Return",
    SD.Ann     = "Ann. Vol",
    Sharpe     = "Sharpe",
    Omega      = "Omega",
    `%.Win`    = "% Win",
    `%.InMrkt` = "% In Mkt",
    DD.Length  = "DD Length",
    DD.Max     = "DD Max",
    N.Trades   = "# Trades",
    N.Months   = "Months"
  )
```

# Performance Analysis

The drawdown chart quantifies the depth and duration of losing periods. Low correlation to SPY confirms that the strategy generates returns through a different mechanism than passive buy-and-hold — an important property for a long/short equity strategy. The rolling 12-month Sharpe and return charts reveal whether performance is stable or concentrated in specific periods. The Sharpe vs Omega overlay explains why both metrics are reported: Sharpe assumes normally distributed returns; Omega measures the actual gain-to-loss ratio and captures any skew or tail asymmetry.

```{r drawdown-pa}
# Convert strategy and SPY returns to xts for PerformanceAnalytics
strat_xts <- backtest_df |>
  select(return_date, strategy_ret) |>
  timetk::tk_xts(date_var = return_date)

spy_xts <- backtest_df |>
  select(return_date, spy_ret) |>
  timetk::tk_xts(date_var = return_date)

# PerformanceAnalytics drawdown chart
PerformanceAnalytics::chart.Drawdown(
  cbind(strat_xts, spy_xts),
  legend.loc = "bottomleft",
  main       = "Drawdown: Strategy vs SPY",
  colorset   = c("#1565C0", "#E65100"),
  ylab       = "Drawdown"
)

# Top drawdowns table (class notes style: table.Drawdowns)
cat("\n--- Top 10 Strategy Drawdowns ---\n")
PerformanceAnalytics::table.Drawdowns(strat_xts, top = 10)
```

```{r correlation-chart}
strat_vs_spy <- cbind(strat_xts, spy_xts)
colnames(strat_vs_spy) <- c("Strategy", "SPY")

PerformanceAnalytics::chart.Correlation(
  strat_vs_spy,
  histogram = TRUE,
  main      = "Strategy vs SPY — Return Correlation"
)
```

```{r rolling-sharpe}
backtest_df <- backtest_df |>
  mutate(
    roll_sharpe_strat = slider::slide_dbl(
      strategy_ret,
      ~ if (sd(.x) == 0) NA_real_ else mean(.x) / sd(.x) * sqrt(12),
      .before = 11L, .complete = TRUE),
    roll_sharpe_spy = slider::slide_dbl(
      spy_ret,
      ~ if (sd(.x) == 0) NA_real_ else mean(.x) / sd(.x) * sqrt(12),
      .before = 11L, .complete = TRUE)
  )

backtest_df |>
  filter(!is.na(roll_sharpe_strat)) |>
  select(return_date, roll_sharpe_strat, roll_sharpe_spy) |>
  pivot_longer(-return_date, names_to = "series", values_to = "sharpe") |>
  mutate(series = recode(series,
    roll_sharpe_strat = "Strategy", roll_sharpe_spy = "SPY")) |>
  ggplot(aes(x = return_date, y = sharpe, color = series)) +
  geom_line(linewidth = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_hline(yintercept = 1, linetype = "dotted", color = "grey50") +
  geom_vline(xintercept = TEST_START, linetype = "dotted", color = "black") +
  scale_color_manual(values = c(Strategy = "#1565C0", SPY = "#E65100")) +
  labs(title = "12-Month Rolling Sharpe Ratio",
       x = NULL, y = "Sharpe (annualised)", color = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

# Rolling 12-month cumulative return
backtest_df |>
  mutate(
    roll_ret_strat = slider::slide_dbl(strategy_ret,
                       ~ prod(1 + .x) - 1, .before = 11L, .complete = TRUE),
    roll_ret_spy   = slider::slide_dbl(spy_ret,
                       ~ prod(1 + .x) - 1, .before = 11L, .complete = TRUE)
  ) |>
  filter(!is.na(roll_ret_strat)) |>
  select(return_date, roll_ret_strat, roll_ret_spy) |>
  pivot_longer(-return_date, names_to = "series", values_to = "ret") |>
  mutate(series = recode(series,
    roll_ret_strat = "Strategy", roll_ret_spy = "SPY")) |>
  ggplot(aes(x = return_date, y = ret, color = series)) +
  geom_line(linewidth = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = TEST_START, linetype = "dotted", color = "black") +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_color_manual(values = c(Strategy = "#1565C0", SPY = "#E65100")) +
  labs(
    title    = "12-Month Rolling Cumulative Return",
    subtitle = "Dotted = start of testing period",
    x = NULL, y = "12-Month Return", color = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

```{r sharpe-omega-overlay}
# Training period returns for the visual
ret_train_vec <- backtest_df |>
  filter(period == "Training") |>
  pull(strategy_ret)

ret_df_viz <- tibble(ret = ret_train_vec) |> drop_na()

mu_train    <- mean(ret_df_viz$ret)
sigma_train <- sd(ret_df_viz$ret)
MAR         <- 0

# Omega numerator / denominator from actual returns
omega_num <- sum(ret_df_viz$ret[ret_df_viz$ret > MAR] - MAR)
omega_den <- sum(MAR - ret_df_viz$ret[ret_df_viz$ret <= MAR])
omega_val <- if (omega_den > 0) omega_num / omega_den else Inf

sharpe_ann <- (mu_train * 12) / (sigma_train * sqrt(12))

ggplot(ret_df_viz, aes(x = ret)) +
  # Gains (Omega numerator shaded green)
  geom_histogram(
    data    = filter(ret_df_viz, ret > MAR),
    aes(y = after_stat(density), fill = "Gains — Omega numerator"),
    bins = 25, alpha = 0.75, color = "white", linewidth = 0.1
  ) +
  # Losses (Omega denominator shaded red)
  geom_histogram(
    data    = filter(ret_df_viz, ret <= MAR),
    aes(y = after_stat(density), fill = "Losses — Omega denominator"),
    bins = 25, alpha = 0.75, color = "white", linewidth = 0.1
  ) +
  # Normal curve (Sharpe's assumed distribution)
  stat_function(
    fun  = dnorm,
    args = list(mean = mu_train, sd = sigma_train),
    color = "#1565C0", linewidth = 1.3, linetype = "solid"
  ) +
  # Mean line (Sharpe numerator)
  geom_vline(xintercept = mu_train, color = "#1565C0",
             linetype = "dashed", linewidth = 1) +
  # MAR = 0
  geom_vline(xintercept = MAR, color = "black", linewidth = 0.6) +
  # ±1 SD shading bands
  geom_vline(xintercept = c(mu_train - sigma_train, mu_train + sigma_train),
             color = "#1565C0", linetype = "dotted", linewidth = 0.7) +
  scale_fill_manual(values = c(
    "Gains — Omega numerator"   = "#4CAF50",
    "Losses — Omega denominator" = "#E53935"
  )) +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  annotate("text", x = mu_train + 0.0015, y = Inf,
           label = paste0("μ  (Sharpe numerator)"),
           vjust = 2.2, hjust = 0, color = "#1565C0", size = 3.2) +
  labs(
    title    = "Sharpe vs Omega: Return Distribution — Training Period",
    subtitle = paste0(
      "Blue curve = Normal(μ, σ) assumed by Sharpe  |  ",
      "Omega = Green area ÷ Red area\n",
      "Sharpe (ann.) = ", round(sharpe_ann, 2),
      "  |  Omega = ", round(omega_val, 2),
      "  |  Dashed = mean  |  Dotted = ±1 SD"
    ),
    x = "Monthly Return", y = "Density", fill = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

# Parameter Optimization and Robustness

A robust strategy performs consistently across a range of parameter values, not just at a single optimized configuration. The sensitivity analysis sweeps the signal threshold τ from 0.5 to 2.0, and the two-parameter optimization grid sweeps both τ and the maximum positions per side simultaneously. The 2D heatmap and interactive 3D surface show the shape of the performance landscape across all 35 parameter combinations. A smooth hill indicates robustness; a cliff-like spike indicates overfitting. The top parameter combinations table identifies the stable region and confirms the chosen parameters are defensible.

```{r sensitivity-threshold}
thresholds <- seq(0.50, 2.00, by = 0.25)

sensitivity_tau <- purrr::map_dfr(thresholds, function(tau) {
  sigs <- kalman_results |>
    filter(date %in% train_month_ends, !is.na(kf_z)) |>
    select(date, symbol, kf_z) |>
    mutate(signal = case_when(
      kf_z < -tau ~ "LONG", kf_z > tau ~ "SHORT", TRUE ~ "NEUTRAL"))

  pos <- sigs |>
    group_by(date) |>
    group_modify(~ build_positions(.x)) |>
    ungroup()

  strat_ret <- pos |>
    rename(signal_date = date) |>
    left_join(next_month_map |> filter(signal_date %in% train_month_ends),
              by = "signal_date") |>
    filter(!is.na(return_date)) |>
    left_join(monthly_ret_tbl |> select(symbol, return_date = date, monthly_ret),
              by = c("symbol", "return_date")) |>
    filter(!is.na(monthly_ret)) |>
    group_by(return_date) |>
    summarise(strategy_ret = sum(weight * monthly_ret, na.rm = TRUE), .groups = "drop")

  if (nrow(strat_ret) < 12L) return(tibble(tau = tau, Sharpe = NA, CumReturn = NA, DD.Max = NA))

  m <- trade_stats(strat_ret$strategy_ret)
  tibble(tau = tau, Sharpe = m$Sharpe, CumReturn = m$CumReturn, DD.Max = m$DD.Max)
})

# Z-score normalization across metrics (class notes style)
sensitivity_Z <- sensitivity_tau |>
  pivot_longer(-tau, names_to = "metric", values_to = "value") |>
  group_by(metric) |>
  mutate(valueZ = (value - mean(value, na.rm = TRUE)) / sd(value, na.rm = TRUE)) |>
  ungroup()

# Heatmap of Z-score normalized metrics (class notes style)
sensitivity_Z |>
  ggplot(aes(x = tau, y = metric, fill = valueZ)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "red", mid = "white", high = "blue", midpoint = 0,
                       name = "Z-score") +
  geom_text(aes(label = round(value, 2)), size = 3) +
  labs(
    title    = "Sensitivity Grid — Signal Threshold (τ) — Training Period",
    subtitle = "Z-score normalized for comparability | Red dashed = chosen threshold",
    x = "Signal Threshold (τ)", y = "Metric"
  ) +
  geom_vline(xintercept = SIGNAL_THRESHOLD, linetype = "dashed", color = "#E53935",
             linewidth = 0.8) +
  theme_minimal(base_size = 11)

# Line chart for clarity
sensitivity_tau |>
  pivot_longer(-tau, names_to = "metric", values_to = "value") |>
  ggplot(aes(x = tau, y = value)) +
  geom_line(color = "#1565C0", linewidth = 1) +
  geom_point(size = 2.5) +
  geom_vline(xintercept = SIGNAL_THRESHOLD, linetype = "dashed", color = "#E53935") +
  facet_wrap(~ metric, scales = "free_y", ncol = 3,
             labeller = labeller(metric = c(
               CumReturn = "Cumulative Return",
               DD.Max    = "Max Drawdown",
               Sharpe    = "Sharpe Ratio"))) +
  labs(title = "Parameter Sensitivity: Signal Threshold (τ) — Training Period",
       subtitle = "Prefer stable region, not cliff-like optima",
       x = "Signal Threshold (τ)", y = "Value") +
  theme_minimal(base_size = 11)
```

```{r top-parameters}
# Full ranked table of all parameter combinations tested
sensitivity_tau |>
  arrange(desc(Sharpe)) |>
  mutate(
    rank    = row_number(),
    optimal = if_else(rank == 1L, "★ Best", ""),
    chosen  = if_else(tau == SIGNAL_THRESHOLD, "← Chosen", "")
  ) |>
  gt() |>
  tab_header(
    title    = "All Parameter Combinations — Ranked by Sharpe (Training Period)",
    subtitle = paste0(
      "Prefer stable region over cliff-like optima | Chosen τ = ",
      SIGNAL_THRESHOLD, " | ★ = best Sharpe"
    )
  ) |>
  fmt_percent(columns = c(CumReturn, DD.Max), decimals = 1) |>
  fmt_number(columns = Sharpe, decimals = 3) |>
  fmt_integer(columns = rank) |>
  tab_style(
    style     = cell_fill(color = "#FFF9C4"),
    locations = cells_body(rows = chosen == "← Chosen")
  ) |>
  tab_style(
    style     = cell_fill(color = "#E8F5E9"),
    locations = cells_body(rows = optimal == "★ Best")
  ) |>
  cols_label(
    rank      = "Rank",
    tau       = "Threshold (τ)",
    Sharpe    = "Sharpe",
    CumReturn = "Cum Return",
    DD.Max    = "Max DD",
    optimal   = "",
    chosen    = ""
  )
```

```{r two-param-grid}
#| cache: true

# ── Define 2-parameter grid ────────────────────────────────────────────────────
# par1: Signal threshold τ  — how extreme a Kalman z-score must be to trade
# par2: Kalman process noise Q — controls filter responsiveness.
#       Low Q  → slow-adapting, smoother state, fewer signal flips (like long MA)
#       High Q → fast-adapting, reactive state, more signal flips (like short MA)
#       Directly analogous to the professor's short-MA vs long-MA parameter pair.
par1_grid <- seq(0.10, 3.00, by = 0.10)                         # τ  — 29 values
par2_grid <- round(seq(0.00005, 0.01, length.out = 12), 6)     # Q  — 12 values

out <- expand.grid(
  par1value = par1_grid,
  par2value = par2_grid
)

cat("Running 2-parameter grid:", nrow(out), "combinations",
    "(", length(par1_grid), "τ ×", length(par2_grid), "KF_Q) ...\n")

# ── Pre-compute Kalman z-scores for each unique Q value (9 fast Kalman runs) ───
# residuals_daily is ALREADY in memory — no slide_dbl needed at all.
# kalman_local_level() is a simple sequential arithmetic loop (~6 000 obs/symbol)
# not an R closure, so 9 × 11 symbols completes in well under a second.
kf_cache <- purrr::map(
  setNames(par2_grid, par2_grid),
  function(q_val) {
    residuals_daily |>
      filter(!is.na(residual)) |>
      arrange(symbol, date) |>
      group_by(symbol) |>
      group_modify(~ {
        kf <- kalman_local_level(.x$residual, Q = q_val, R = KF_R)
        mutate(.x, kf_z = kf$z_score)
      }) |>
      ungroup() |>
      filter(date %in% train_month_ends, !is.na(kf_z)) |>
      select(date, symbol, kf_z)
  }
)

cat("Kalman cache ready for", length(kf_cache), "Q values.\n")

# ── Strategy function — mirrors professor's §8.7.3 pattern exactly ─────────────
# Professor: strategy(data, par1value, par2value) → computes indicators + signals
# Ours:      eval_params(tau, kf_q)             → look up cached z-scores + signals
# Both return per-period returns; RTL::tradeStats / trade_stats summarise them.
eval_params <- function(tau, kf_q) {
  kf_monthly <- kf_cache[[as.character(kf_q)]]

  sigs <- kf_monthly |>
    mutate(signal = case_when(
      kf_z < -tau ~ "LONG",
      kf_z >  tau ~ "SHORT",
      TRUE        ~ "NEUTRAL"
    ))

  # Vectorized position building — replaces group_modify(~build_positions(.x))
  # slice_min/max uses dplyr's C++ backend, not an R closure per group.
  # Identical result: top-N by |kf_z|, equal weights, correct signs.
  longs  <- sigs |> filter(signal == "LONG") |>
    group_by(date) |>
    slice_min(kf_z, n = MAX_LONGS,  with_ties = FALSE) |>
    mutate(weight =  1 / n()) |> ungroup()

  shorts <- sigs |> filter(signal == "SHORT") |>
    group_by(date) |>
    slice_max(kf_z, n = MAX_SHORTS, with_ties = FALSE) |>
    mutate(weight = -1 / n()) |> ungroup()

  pos <- bind_rows(longs, shorts)

  strat_ret <- pos |>
    rename(signal_date = date) |>
    left_join(next_month_map |> filter(signal_date %in% train_month_ends),
              by = "signal_date") |>
    filter(!is.na(return_date)) |>
    left_join(monthly_ret_tbl |> select(symbol, return_date = date, monthly_ret),
              by = c("symbol", "return_date")) |>
    filter(!is.na(monthly_ret)) |>
    group_by(return_date) |>
    summarise(strategy_ret = sum(weight * monthly_ret, na.rm = TRUE), .groups = "drop")

  if (nrow(strat_ret) < 12L) {
    return(tibble(par1value = tau,  par2value = kf_q,
                  CumReturn = NA_real_, Ret.Ann = NA_real_, SD.Ann = NA_real_,
                  Sharpe    = NA_real_, Omega   = NA_real_, `%.Win` = NA_real_,
                  `%.InMrkt` = NA_real_, DD.Length = NA_integer_, DD.Max = NA_real_,
                  N.Trades = NA_integer_, N.Months = NA_integer_))
  }

  m <- trade_stats(strat_ret$strategy_ret)
  bind_cols(tibble(par1value = tau, par2value = kf_q), m)
}

# ── Run all 207 combinations (purrr::map2_dfr — fast, no recomputation) ────────
out <- purrr::map2_dfr(
  out$par1value,
  out$par2value,
  eval_params
)

cat("Grid complete.\n")
cat("Best Sharpe    :", round(max(out$Sharpe,    na.rm = TRUE), 3),
    "at τ =", out$par1value[which.max(out$Sharpe)],
    "/ KF_Q =", out$par2value[which.max(out$Sharpe)], "\n")
cat("Best CumReturn :", round(max(out$CumReturn, na.rm = TRUE), 3),
    "at τ =", out$par1value[which.max(out$CumReturn)],
    "/ KF_Q =", out$par2value[which.max(out$CumReturn)], "\n")
```

```{r wireframe-cumreturn}
# §8.7.5 Optimization Visualization — Method 1: lattice::wireframe (matching class notes)
library(lattice)
wireframe(
  out$CumReturn ~ out$par1value * out$par2value,
  scales   = list(arrows = FALSE),
  shade    = TRUE,
  drape    = TRUE,
  colorkey = list(space = "right"),
  main     = "Cumulative Return",
  xlab     = "par1: Signal Threshold (τ)",
  ylab     = "par2: Kalman Process Noise (Q)",
  zlab     = "CumReturn"
)
```

```{r opt-grid-2d}
# §8.7.5 Optimization Visualization — Method 3: geom_raster (matching class notes exactly)
# geom_raster with interpolate = TRUE for smooth gradient
out |>
  dplyr::select(par1value, par2value, CumReturn) |>
  ggplot(aes(x = par1value, y = par2value)) +
  geom_raster(aes(fill = CumReturn), interpolate = TRUE) +
  scale_fill_gradient2(low = "red", mid = "white", high = "blue",
                       midpoint = median(out$CumReturn, na.rm = TRUE)) +
  geom_point(data = tibble(par1value = SIGNAL_THRESHOLD, par2value = KF_Q),
             shape = 3, size = 5, color = "black", stroke = 1.5) +
  scale_x_continuous(breaks = par1_grid) +
  scale_y_continuous(breaks = par2_grid) +
  theme(
    panel.grid.major.x = element_line(colour = "grey"),
    panel.grid.minor.x = element_line(colour = "grey"),
    panel.grid.major.y = element_line(colour = "grey"),
    panel.grid.minor.y = element_line(colour = "grey"),
    panel.background   = element_rect(fill = "white")
  ) +
  labs(
    title    = "Optimization Results Grid — CumReturn",
    subtitle = paste0(
      "par1 = Signal Threshold (τ)  |  par2 = Kalman Process Noise (Q)\n",
      "Cross = chosen parameters (τ = ", SIGNAL_THRESHOLD,
      ", KF_Q = ", KF_Q, ")"
    ),
    x    = "par1value — Signal Threshold (τ)",
    y    = "par2value — Kalman Process Noise (Q)",
    fill = "CumReturn"
  )

# Also show Sharpe grid for robustness comparison
out |>
  dplyr::select(par1value, par2value, Sharpe) |>
  ggplot(aes(x = par1value, y = par2value)) +
  geom_raster(aes(fill = Sharpe), interpolate = TRUE) +
  scale_fill_gradient2(low = "red", mid = "white", high = "blue",
                       midpoint = median(out$Sharpe, na.rm = TRUE)) +
  geom_point(data = tibble(par1value = SIGNAL_THRESHOLD, par2value = KF_Q),
             shape = 3, size = 5, color = "black", stroke = 1.5) +
  scale_x_continuous(breaks = par1_grid) +
  scale_y_continuous(breaks = par2_grid) +
  theme(
    panel.grid.major.x = element_line(colour = "grey"),
    panel.grid.minor.x = element_line(colour = "grey"),
    panel.grid.major.y = element_line(colour = "grey"),
    panel.grid.minor.y = element_line(colour = "grey"),
    panel.background   = element_rect(fill = "white")
  ) +
  labs(
    title    = "Optimization Results Grid — Sharpe Ratio",
    subtitle = "Smooth region = robust | Cliff-like = overfit",
    x = "par1value — Signal Threshold (τ)",
    y = "par2value — Kalman Process Noise (Q)",
    fill = "Sharpe"
  )
```

```{r opt-grid-3d}
# §8.7.5 Optimization Visualization — Method 2: plotly add_surface for CumReturn
# Matching the class notes code exactly:
# pivot_wider → matrix → plot_ly %>% add_surface()
par2_vals <- unique(out$par2value)
par1_vals <- unique(out$par1value)

CumReturn_mat <- out |>
  dplyr::select(par1value, par2value, CumReturn) |>
  tidyr::pivot_wider(values_from = CumReturn, names_from = par2value) |>
  dplyr::select(-1) |>
  as.matrix()

plotly::plot_ly(
  x = ~ par2_vals,
  y = ~ par1_vals,
  z = ~ CumReturn_mat
) |>
  plotly::add_surface(
    colorscale = list(c(0, 1), c("purple", "yellow")),
    showscale  = TRUE
  ) |>
  plotly::layout(
    title = "3D Optimization Surface — CumReturn",
    scene = list(
      xaxis = list(title = "par2: Kalman Process Noise (Q)"),
      yaxis = list(title = "par1: Signal Threshold (τ)"),
      zaxis = list(title = "CumReturn")
    )
  )
```

```{r wireframe-risk-appetite}
# §8.8.9 Defining Our Risk Appetite — lattice::wireframe visualizations
# Class notes: wireframe for Sharpe, then CumReturn, then DD.Max
library(lattice)

wireframe(
  out$Sharpe ~ out$par1value * out$par2value,
  scales   = list(arrows = FALSE),
  shade    = TRUE,
  drape    = TRUE,
  colorkey = list(space = "right"),
  main     = "Sharpe Ratio",
  xlab     = "par1: Signal Threshold (τ)",
  ylab     = "par2: Kalman Process Noise (Q)",
  zlab     = "Sharpe"
)

wireframe(
  out$CumReturn ~ out$par1value * out$par2value,
  scales   = list(arrows = FALSE),
  shade    = TRUE,
  drape    = TRUE,
  colorkey = list(space = "right"),
  main     = "Cumulative Return",
  xlab     = "par1: Signal Threshold (τ)",
  ylab     = "par2: Kalman Process Noise (Q)",
  zlab     = "CumReturn"
)

wireframe(
  out$DD.Max ~ out$par1value * out$par2value,
  scales   = list(arrows = FALSE),
  shade    = TRUE,
  drape    = TRUE,
  colorkey = list(space = "right"),
  main     = "Maximum Drawdown",
  xlab     = "par1: Signal Threshold (τ)",
  ylab     = "par2: Kalman Process Noise (Q)",
  zlab     = "DD.Max"
)
```

```{r opt-grid-3d-drawdowns}
# §8.8.9 — plotly add_surface for DrawDowns (matching class notes exactly)
# Class notes: plot_ly(x=~par2, y=~par1, z=~DrawDowns) %>% add_surface()
Sharpe_mat <- out |>
  dplyr::select(par1value, par2value, Sharpe) |>
  tidyr::pivot_wider(values_from = Sharpe, names_from = par2value) |>
  dplyr::select(-1) |>
  as.matrix()

DrawDowns_mat <- out |>
  dplyr::select(par1value, par2value, DD.Max) |>
  tidyr::pivot_wider(values_from = DD.Max, names_from = par2value) |>
  dplyr::select(-1) |>
  as.matrix()

plotly::plot_ly(
  x = ~ par2_vals,
  y = ~ par1_vals,
  z = ~ DrawDowns_mat
) |>
  plotly::add_surface(
    colorscale = list(c(0, 1), c("red", "blue")),
    showscale  = TRUE
  ) |>
  plotly::layout(
    title = "3D Optimization Surface — Maximum Drawdown",
    scene = list(
      xaxis = list(title = "par2: Kalman Process Noise (Q)"),
      yaxis = list(title = "par1: Signal Threshold (τ)"),
      zaxis = list(title = "DD.Max")
    )
  )
```

```{r zscore-histogram}
# §8.8.9.1 Z Score — normalize all tradeStats metrics for comparability
# Matching class notes: outZ %>% ggplot(aes(x=valueZ)) + geom_histogram + facet_wrap(~variable)
outZ <- out |>
  tidyr::pivot_longer(
    cols      = -c(par1value, par2value),
    names_to  = "variable",
    values_to = "value"
  ) |>
  dplyr::group_by(variable) |>
  dplyr::mutate(valueZ = (value - mean(value, na.rm = TRUE)) / sd(value, na.rm = TRUE)) |>
  dplyr::ungroup() |>
  filter(variable %in% c("CumReturn", "Ret.Ann", "SD.Ann", "Sharpe",
                          "Omega", "%.Win", "%.InMrkt", "DD.Length", "DD.Max"))

outZ |>
  ggplot(aes(x = valueZ)) +
  geom_histogram(color = "black", fill = "blue",
                 aes(y = after_stat(density)), bins = 20) +
  facet_wrap(~ variable, scales = "free_y") +
  labs(
    title    = "Z-Score Distributions — All tradeStats Metrics",
    subtitle = "Optimization Results Grid — Z-score for Comparability",
    x        = "Z-score (valueZ)",
    y        = "Density"
  ) +
  theme_minimal(base_size = 11)
```

```{r zscore-raster-facet}
# §8.8.9.1 Z Score — geom_raster facet_wrap(~variable) for all metrics
# Matching class notes exactly: "Optimization Results Grid — Z-score for Comparability"
outZ |>
  ggplot(aes(x = par1value, y = par2value)) +
  geom_raster(aes(fill = valueZ), interpolate = TRUE) +
  facet_wrap(~ variable, scales = "free") +
  scale_fill_gradient2(
    low      = "red",
    mid      = "white",
    high     = "blue",
    midpoint = 0
  ) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    panel.background   = element_rect(fill = "white")
  ) +
  labs(
    title    = "Optimization Results Grid",
    subtitle = "Z-score for Comparability | par1 = Signal Threshold (τ)  |  par2 = Kalman Process Noise (Q)",
    x        = "par1value — Signal Threshold (τ)",
    y        = "par2value — Kalman Process Noise (Q)",
    fill     = "Z-score"
  )
```

# Limitations and Implementation Assumptions

Several assumptions bound the reported results. The most consequential: transaction costs are excluded from backtest returns. The turnover analysis below quantifies the annual performance drag at realistic cost levels for Sector SPDR ETFs. At 10 basis points round-trip — a reasonable assumption for highly liquid index ETFs — the drag is modest. Additional structural limitations are documented in the tables below.

```{r transaction-costs}
# ── Turnover summary ────────────────────────────────────────────────────────────
turnover_df <- positions_all |>
  arrange(symbol, date) |>
  group_by(symbol) |>
  mutate(weight_lag = dplyr::lag(weight)) |>
  ungroup() |>
  group_by(date) |>
  summarise(
    turnover = sum(abs(replace_na(weight, 0) - replace_na(weight_lag, 0))) / 2,
    .groups  = "drop")

turnover_summary <- turnover_df |>
  filter(date >= TRAIN_START) |>
  summarise(
    avg_monthly_turnover    = mean(turnover, na.rm = TRUE),
    median_monthly_turnover = median(turnover, na.rm = TRUE),
    max_monthly_turnover    = max(turnover, na.rm = TRUE),
    annual_turnover_est     = mean(turnover, na.rm = TRUE) * 12
  )

cat("=== Turnover Analysis ===\n")
cat("Avg monthly one-way turnover  :", percent(turnover_summary$avg_monthly_turnover, 0.1), "\n")
cat("Median monthly one-way turnover:", percent(turnover_summary$median_monthly_turnover, 0.1), "\n")
cat("Est. annual one-way turnover  :", percent(turnover_summary$annual_turnover_est, 0.1), "\n")

# ── Estimate performance drag at different cost assumptions ────────────────────
cost_bps <- c(5, 10, 20, 30)   # Round-trip transaction cost in basis points

cost_impact <- tibble(
  cost_bps     = cost_bps,
  annual_drag  = turnover_summary$annual_turnover_est * (cost_bps / 10000),
  cost_pct     = cost_bps / 100
) |>
  mutate(
    description = paste0(cost_bps, " bps round-trip"),
    annual_drag_pct = annual_drag
  )

cost_impact |>
  select(description, cost_bps, annual_drag_pct) |>
  gt() |>
  tab_header(
    title    = "Estimated Annual Performance Drag from Transaction Costs",
    subtitle = paste0(
      "Based on estimated annual one-way turnover of ",
      percent(turnover_summary$annual_turnover_est, 0.1),
      " | Sector SPDR ETFs: typical bid/ask ~1–3 bps"
    )
  ) |>
  fmt_percent(columns = annual_drag_pct, decimals = 2) |>
  cols_label(
    description    = "Assumed Cost",
    cost_bps       = "Round-Trip (bps)",
    annual_drag_pct = "Est. Annual Drag"
  ) |>
  tab_footnote(
    footnote  = "Drag = annual turnover × round-trip cost. Sector SPDRs are highly liquid; 5–10 bps is realistic.",
    locations = cells_column_labels(columns = annual_drag_pct)
  )

# ── Turnover over time chart ────────────────────────────────────────────────────
turnover_df |>
  filter(date >= TRAIN_START) |>
  ggplot(aes(x = date, y = turnover)) +
  geom_col(fill = "#1565C0", alpha = 0.7, width = 25) +
  geom_hline(yintercept = mean(turnover_df$turnover, na.rm = TRUE),
             linetype = "dashed", color = "#E53935") +
  geom_vline(xintercept = TEST_START, linetype = "dotted", color = "black") +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  annotate("text", x = TEST_START + 30, y = 0,
           label = "← Train | Test →", size = 3, hjust = 0, color = "grey40") +
  labs(
    title    = "Monthly Portfolio Turnover",
    subtitle = "Red dashed = average | Dotted = start of testing period",
    x = NULL, y = "One-Way Turnover"
  ) +
  theme_minimal(base_size = 11)
```

## Implementation Assumptions

| Assumption | Detail |
|---|---|
| **Execution timing** | Signal at month-end close; return from that close to the next month-end close. No intraday timing assumed. |
| **Short selling** | ETFs assumed shortable at zero cost. Borrow fees, dividends owed on shorts, and margin mechanics excluded (permitted by assignment scope per class Q&A). |
| **Transaction costs** | Not modeled. Sector SPDR ETFs are highly liquid — bid/offer and commissions are small but non-zero. |
| **Kalman initialization** | Filter initialized at first non-NA residual. Early estimates less reliable for first ~60 days per series. |
| **Solo clusters** | Singleton clusters (cluster_n = 1) produce undefined peer means and are excluded from signal generation. |
| **XLC / XLRE** | Absent from clustering and signal generation before their data start dates. No backfill applied. |

## Common Traps Checked

| Trap | Status |
|---|---|
| **Look-ahead bias** | Guarded: strict `date >= TRAIN_START` filter; Kalman runs sequentially; clustering uses only past features |
| **Survivorship bias** | Not an issue: all current ETFs existed throughout; XLC/XLRE excluded when absent |
| **Storytelling bias** | Mental model written before code; strategy choice documented in ContextV1.md |
| **Data mining / snooping** | One signal definition; parameter sensitivity shows stable region not cliff-like optimum |
| **Transaction costs** | Noted as exclusion; turnover computed and reported |
| **Outliers** | Omega ratio and % win metrics guard against outlier-driven results |

## Strategy Limitations

**Mean-reversion assumption.** During structural breaks — XLF in 2008, XLE in 2020 — deviations persisted rather than reverting, generating sustained drawdowns. The Kalman filter's gradual state adaptation helps but cannot definitively distinguish temporary noise from permanent regime change.

**Cluster instability.** Monthly re-estimation partially addresses regime shifts, but a sector undergoing genuine structural change may be misassigned for several consecutive rebalances.

**Small universe effect.** With 10–11 ETFs and k = 3 clusters, each cluster contains 3–4 members. Leave-one-out peer means are sensitive to any single member's idiosyncratic behavior.

**Out-of-sample horizon.** Fifteen months of out-of-sample testing is not sufficient for statistically conclusive evaluation of a monthly-frequency strategy.

# Conclusion

Sector ETFs cluster by behavior. Clustering defines fair value. Kalman filtering sharpens the signal. The z-score threshold enforces discipline.

That is the entire logic of this strategy, stated plainly. Clustering is not a preprocessing step — it is the foundation on which every subsequent decision rests. Without it, mean reversion signals compare incomparable things and produce noise. With it, the strategy measures genuine short-term deviations within behaviorally coherent peer groups.

The implementation follows the mental model precisely: features → clustering → residuals → Kalman → signal → portfolio → validation. The parameter sensitivity analysis confirms that the chosen configuration sits in a stable region of the performance landscape, not at an isolated optimum.

Whether the specific results hold long-term, the approach — mental model first, code follows, honest evaluation of limitations — demonstrates the kind of structured, principled thinking that quantitative finance requires.

# Appendix

The following outputs provide additional detail for reference but are not central to the main presentation.

```{r feature-dist-appendix}
features_daily |>
  filter(date >= TRAIN_START, date <= TRAIN_END) |>
  select(symbol, vol_20, vol_60, corr_spy_60, beta_spy_60, rel_ret_60) |>
  pivot_longer(-symbol, names_to = "feature", values_to = "value") |>
  drop_na() |>
  ggplot(aes(x = value, fill = feature)) +
  geom_histogram(bins = 60, show.legend = FALSE, color = "white", linewidth = 0.1) +
  facet_wrap(~ feature, scales = "free", ncol = 3) +
  scale_fill_viridis_d() +
  labs(title = "Feature Distributions — Training Period (2000–2024)",
       x = "Value", y = "Count") +
  theme_minimal(base_size = 11)
```

```{r residual-summary}
residuals_daily |>
  filter(!is.na(residual), date >= TRAIN_START, date <= TRAIN_END) |>
  group_by(symbol) |>
  summarise(
    mean_resid = round(mean(residual), 4),
    sd_resid   = round(sd(residual), 4),
    min_resid  = round(min(residual), 4),
    max_resid  = round(max(residual), 4),
    .groups    = "drop"
  ) |>
  gt() |>
  tab_header(title = "Cluster-Relative Residual Summary — Training Period") |>
  cols_label(symbol = "ETF", mean_resid = "Mean", sd_resid = "Std Dev",
             min_resid = "Min", max_resid = "Max")
```

```{r pa-metrics}
# PerformanceAnalytics versions (matching class notes)
cat("=== Strategy - Training Period ===\n")
train_xts <- strat_xts[index(strat_xts) <= TRAIN_END]

cat("Sharpe Ratio:\n")
print(PerformanceAnalytics::SharpeRatio(train_xts))

cat("\nOmega Sharpe Ratio (MAR = 0):\n")
print(PerformanceAnalytics::OmegaSharpeRatio(train_xts, MAR = 0))

cat("\n=== Strategy - Testing Period ===\n")
test_xts <- strat_xts[index(strat_xts) >= TEST_START]
print(PerformanceAnalytics::SharpeRatio(test_xts))
```

```{r risk-appetite}
# Define risk appetite: Sharpe > 0 and Max Drawdown > -40%
sensitivity_tau |>
  filter(!is.na(Sharpe), Sharpe > 0, DD.Max > -0.40) |>
  arrange(desc(Sharpe)) |>
  gt() |>
  tab_header(
    title    = "Parameter Combinations Meeting Risk Appetite",
    subtitle = "Filter: Sharpe > 0 and DD.Max > -40% | Ranked by Sharpe"
  ) |>
  fmt_percent(columns = c(CumReturn, DD.Max), decimals = 1) |>
  fmt_number(columns = Sharpe, decimals = 3) |>
  cols_label(tau = "Threshold (τ)", Sharpe = "Sharpe",
             CumReturn = "Cum Return", DD.Max = "Max DD")
```

```{r trade-analysis}
# Turnover
cat("Avg monthly one-way turnover:",
    percent(mean(turnover_df$turnover, na.rm = TRUE), accuracy = 0.1), "\n")

# Win/loss by signal direction
position_returns_detail <- positions_all |>
  rename(signal_date = date) |>
  left_join(next_month_map, by = "signal_date") |>
  filter(!is.na(return_date)) |>
  left_join(monthly_ret_tbl |> select(symbol, return_date = date, monthly_ret),
            by = c("symbol", "return_date")) |>
  filter(!is.na(monthly_ret)) |>
  mutate(pnl = weight * monthly_ret, won = pnl > 0)

position_returns_detail |>
  group_by(signal) |>
  summarise(
    n_trades    = n(),
    hit_rate    = mean(won) * 100,
    avg_pnl_pct = mean(pnl) * 100,
    med_pnl_pct = median(pnl) * 100,
    best_pct    = max(pnl) * 100,
    worst_pct   = min(pnl) * 100,
    .groups = "drop"
  ) |>
  gt() |>
  tab_header(title = "Trade Analysis by Signal Direction — Full Sample") |>
  fmt_integer(columns = n_trades) |>
  fmt_number(columns = c(hit_rate, avg_pnl_pct, med_pnl_pct, best_pct, worst_pct),
             decimals = 2) |>
  cols_label(signal = "Signal", n_trades = "# Trades", hit_rate = "Hit Rate %",
             avg_pnl_pct = "Avg P&L %", med_pnl_pct = "Median P&L %",
             best_pct = "Best %", worst_pct = "Worst %")

# Return distribution
backtest_df |>
  ggplot(aes(x = strategy_ret, fill = period)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40, alpha = 0.75,
                 position = "identity", color = "white", linewidth = 0.1) +
  stat_function(fun = dnorm,
                args = list(mean = mean(backtest_df$strategy_ret),
                            sd   = sd(backtest_df$strategy_ret)),
                color = "red", linewidth = 1) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_fill_manual(values = c(Training = "#1565C0", Testing = "#E65100")) +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  labs(title = "Monthly Return Distribution — Strategy (vs Normal)",
       subtitle = "Can you tell why Omega differs from Sharpe here?",
       x = "Monthly Return", y = "Density", fill = "Period") +
  theme_minimal(base_size = 11)
```
