---
title: "Cluster-Relative Mean Reversion on S&P 500 Sector ETFs"
subtitle: "FIN452 — Quantitative Trading Final Project"
author: "Will Thompson"
date: "`r Sys.Date()`"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    toc-float: true
    theme: flatly
    code-fold: true
    code-summary: "Show Code"
    fig-width: 10
    fig-height: 6
execute:
  warning: false
  message: false
  cache: true
  echo: true
---

# Mental Model

## Strategy Overview

This project implements a **cluster-relative mean reversion** strategy on the S&P 500 Select Sector SPDR ETFs. The core hypothesis is that sector ETFs form behavioral clusters, and temporary deviations of an ETF from its cluster-relative fair value contain a tradable mean-reversion signal.

**Clustering is the primary signal engine.** It determines which ETFs are comparable at each rebalance date and defines what "fair value" means for each ETF. The **Kalman filter refines** — not replaces — the cluster-relative signal by smoothing the residual series and providing a statistically interpretable z-score for trade entry.

## Mental Model Diagram

```{mermaid}
%%| fig-width: 10
%%| fig-height: 16
flowchart TD
    subgraph P["① Problem Framing"]
        P1["Can temporary divergences among sector ETFs\nbe traded when sectors are first grouped\ninto behaviorally similar clusters?"]
        P2["Constraints: Daily data · Monthly rebalance\nSPY + Select Sector ETFs · R / Quarto\nClustering primary · Kalman required\nCode must match mental model"]
    end

    subgraph MM["② Mental Model Toolbox"]
        M1["Systems Thinking\nETFs are interconnected,\nnot isolated assets"]
        M2["Inversion\nWhat causes false signals\nor model failure?"]
        M3["Margin of Safety\nOnly trade sufficiently\nlarge deviations"]
        M4["Second-Order Thinking\nDistinguish temporary noise\nfrom structural change"]
        M5["Risk–Reward Framework\nSignal must justify risk,\ndrawdown, and turnover"]
        M6["Opportunity Cost\nCapital to strongest\nsignals only"]
    end

    subgraph MV["③ Market View / Economic Logic"]
        MV1["Sectors form behavioral clusters\nbased on rolling return patterns"]
        MV2["Within a cluster, one ETF can temporarily\novershoot or undershoot peers"]
        MV3["Fair value is cluster-relative,\nnot global — apples to apples only"]
    end

    subgraph DI["④ Daily Inputs"]
        D1["Adjusted daily prices & returns"]
        D2["Rolling 20-day & 60-day volatility"]
        D3["Rolling 60-day correlation to SPY"]
        D4["Rolling 60-day relative return vs SPY"]
        D5["Rolling 60-day beta to SPY"]
    end

    subgraph CL["⑤ PRIMARY SIGNAL ENGINE: Clustering"]
        C1["Elbow method → choose k clusters"]
        C2["K-Means + Hierarchical (Ward.D2)\nAgreement = robust structure"]
        C3["Silhouette validation\nSilhouette > 0.5 = well-clustered"]
        C4["Fair value is cluster-relative, not global"]
    end

    subgraph RE["⑥ Residual Construction"]
        R1["60-day compounded return for each ETF"]
        R2["Cluster benchmark =\nleave-one-out peer mean return"]
        R3["Residual = ETF cum return − cluster benchmark"]
    end

    subgraph KF["⑦ Kalman Filter — Smoothed State Estimate"]
        K1["Local-level state-space model\napplied to each ETF's residual series"]
        K2["Estimate latent fair-value equilibrium\nsmooth noise, improve reliability"]
        K3["Standardized innovation = z-score signal\nKalman refines, does not replace clustering"]
    end

    subgraph DR["⑧ Decision Rules"]
        DR1{"|z-score| ≥ threshold τ?"}
        DR2["z < −τ → LONG\nETF below Kalman fair value\nExpect reversion upward"]
        DR3["z > +τ → SHORT\nETF above Kalman fair value\nExpect reversion downward"]
        DR4["|z| < τ → NEUTRAL\nDeviation too small\nMargin of safety not met"]
        DR5["Rank signals by |z|\nHold only strongest N per side"]
    end

    subgraph PC["⑨ Portfolio Construction"]
        PC1["Monthly rebalance at month-end close"]
        PC2["Equal-weight long & short positions"]
        PC3["Signal threshold = margin of safety\nMax positions = opportunity cost filter"]
    end

    subgraph VA["⑩ Validation & Iteration"]
        V1["Training 2000–2024: parameter selection"]
        V2["Testing 2025–2026: out-of-sample evaluation"]
        V3["tradeStats: Sharpe, Omega, %.Win,\n%.InMrkt, DD.Max, DD.Length"]
        V4["Bias checks: look-ahead, overfitting,\nstorytelling, structural breaks"]
    end

    P --> MM --> MV --> DI --> CL --> RE --> KF --> DR
    DR --> DR1
    DR1 --> DR2
    DR1 --> DR3
    DR1 --> DR4
    DR2 & DR3 --> DR5 --> PC --> VA
    VA -->|"Reflect & refine parameters"| CL
```

---

# 1. Setup

## 1.1 Packages

```{r setup}
#| cache: false

# ── Core tidyverse ─────────────────────────────────────────────────────────────
library(tidyverse)
library(lubridate)

# ── Market data ────────────────────────────────────────────────────────────────
library(tidyquant)
library(timetk)
library(quantmod)

# ── Rolling statistics ─────────────────────────────────────────────────────────
library(slider)
library(TTR)

# ── Tidymodels (preprocessing + data splitting) ────────────────────────────────
library(tidymodels)
library(rsample)
library(recipes)

# ── Clustering ─────────────────────────────────────────────────────────────────
library(cluster)      # silhouette()
library(factoextra)   # fviz_dend(), fviz_silhouette()

# ── Performance analytics ──────────────────────────────────────────────────────
library(PerformanceAnalytics)   # SharpeRatio, Omega, chart.Drawdown, table.Drawdowns

# ── Visualization ──────────────────────────────────────────────────────────────
library(patchwork)
library(scales)
library(viridis)
library(plotly)

# ── Tables ─────────────────────────────────────────────────────────────────────
library(gt)
library(knitr)

# Resolve namespace conflicts
tidymodels_prefer()
```

## 1.2 Strategy Parameters

All unresolved design decisions are centralized here with rationale. Changing any value propagates through the entire document.

```{r parameters}
#| cache: false

# ── Date ranges ────────────────────────────────────────────────────────────────
DATA_START  <- as.Date("2000-01-01")   # Extra history for rolling-feature warmup
TRAIN_START <- as.Date("2000-06-01")   # First full month of training period
TRAIN_END   <- as.Date("2024-12-31")   # End of training period
TEST_START  <- as.Date("2025-01-01")   # Out-of-sample begins
TEST_END    <- as.Date("2026-03-31")   # Inclusive through March 2026

# ── Feature engineering windows (trading days) ─────────────────────────────────
VOL_SHORT   <- 20L   # ~1-month rolling volatility            [DEFAULT: 20]
VOL_LONG    <- 60L   # ~3-month rolling volatility            [DEFAULT: 60]
CORR_WINDOW <- 60L   # Rolling correlation / beta vs SPY      [DEFAULT: 60]
RET_WINDOW  <- 60L   # Rolling cumulative return for residual [DEFAULT: 60]

# ── Clustering ─────────────────────────────────────────────────────────────────
# N_CLUSTERS: Chosen using elbow method (WSS) in section 6.
# With 10-11 ETFs, k=3 gives ~3-4 members per cluster on average.
# [UNRESOLVED] Validate: run elbow method and silhouette across k = 2..5.
N_CLUSTERS  <- 3L
LINK_METHOD <- "ward.D2"   # Ward linkage: minimises within-cluster variance (preferred for finance)

# Features used for clustering
CLUSTER_FEATURES <- c("vol_20", "vol_60", "corr_spy_60", "beta_spy_60", "rel_ret_60")

# ── Kalman filter ──────────────────────────────────────────────────────────────
# [UNRESOLVED] Q/R ratio controls smoothing: low Q/R = slow adaptation (smoother).
# Default Q/R = 0.1 (moderately smooth). Sensitivity tested in section 13.
KF_Q <- 0.001   # Process noise variance  [DEFAULT: 0.001]
KF_R <- 0.01    # Observation noise variance [DEFAULT: 0.01]

# ── Signal thresholds ──────────────────────────────────────────────────────────
# [UNRESOLVED] Margin of safety: only trade |z| >= this value.
SIGNAL_THRESHOLD <- 1.0   # [DEFAULT: 1.0 ~ 1 standard deviation]

# ── Portfolio construction ─────────────────────────────────────────────────────
# [UNRESOLVED] Balances diversification vs signal dilution.
MAX_LONGS  <- 3L   # Maximum long positions per rebalance  [DEFAULT: 3]
MAX_SHORTS <- 3L   # Maximum short positions per rebalance [DEFAULT: 3]

# ── Execution assumption (fixed) ───────────────────────────────────────────────
# Signal at month-end close → position return = next month close-to-close.
# Strictly causal: no future information used at signal generation time.
```

## 1.3 Universe Definition

```{r universe}
#| cache: false

TICKERS <- c(
  "SPY",   # Benchmark — S&P 500
  "XLB",   # Materials
  "XLC",   # Communication Services (from June 2018)
  "XLE",   # Energy
  "XLF",   # Financials
  "XLI",   # Industrials
  "XLK",   # Technology
  "XLP",   # Consumer Staples
  "XLRE",  # Real Estate (from October 2015)
  "XLV",   # Health Care
  "XLU",   # Utilities
  "XLY"    # Consumer Discretionary
)

SECTOR_TICKERS <- setdiff(TICKERS, "SPY")
cat("Universe: SPY (benchmark) +", length(SECTOR_TICKERS), "sector ETFs\n")
```

---

# 2. Data Acquisition

## 2.1 Download Price Data

All data pulled via `tidyquant::tq_get()` (Yahoo Finance API). No local files used.

```{r data-download}
# tq_get() returns: symbol, date, open, high, low, close, volume, adjusted
# 'adjusted' = split and dividend adjusted close — used for all return calculations

raw_prices <- tq_get(
  TICKERS,
  from = DATA_START,
  to   = TEST_END
)

cat("Download complete.\n")
cat("Date range  :", format(min(raw_prices$date)), "to", format(max(raw_prices$date)), "\n")
cat("Rows        :", nrow(raw_prices), "| Symbols:", n_distinct(raw_prices$symbol), "\n")
```

## 2.2 Data Integrity Checks

```{r data-checks}
# ── Check 1: All tickers present ──────────────────────────────────────────────
missing_tickers <- setdiff(TICKERS, unique(raw_prices$symbol))
if (length(missing_tickers) > 0) {
  stop("Missing tickers: ", paste(missing_tickers, collapse = ", "))
} else {
  cat("CHECK 1 PASS: All", length(TICKERS), "tickers downloaded.\n")
}

# ── Check 2: Date coverage and NA rates ───────────────────────────────────────
coverage_tbl <- raw_prices |>
  group_by(symbol) |>
  summarise(
    first_date = min(date),
    last_date  = max(date),
    n_obs      = n(),
    pct_na_adj = round(mean(is.na(adjusted)) * 100, 2),
    .groups    = "drop"
  ) |>
  arrange(first_date)

coverage_tbl |>
  gt() |>
  tab_header(title = "Data Coverage by Ticker") |>
  fmt_date(columns = c(first_date, last_date), date_style = 2) |>
  fmt_number(columns = pct_na_adj, decimals = 2) |>
  cols_label(
    symbol     = "Ticker",
    first_date = "First Date",
    last_date  = "Last Date",
    n_obs      = "Trading Days",
    pct_na_adj = "% Missing (Adj)"
  )

# ── Check 3: No NA in adjusted for standard tickers in training period ─────────
standard_tickers <- setdiff(TICKERS, c("XLC", "XLRE"))
na_standard <- raw_prices |>
  filter(symbol %in% standard_tickers, date >= TRAIN_START, is.na(adjusted)) |>
  nrow()

if (na_standard > 0) {
  warning("NA values found in adjusted prices for standard tickers in training period!")
} else {
  cat("CHECK 3 PASS: No NA in adjusted prices for standard tickers (training period).\n")
}
```

---

# 3. Preprocessing

## 3.1 Adjusted Prices and Daily Returns

Following the course notes pattern — three return types for precise P&L accounting:

```{r preprocessing}
# Keep adjusted close for return calculations.
# Mirror course notes: retClCl (close-to-close), retOpCl (open-to-close same day),
# retClOp (open vs prior close). Used for precise execution assumptions.

prices <- raw_prices |>
  select(symbol, date, open, close, adjusted) |>
  arrange(symbol, date)

returns_daily <- prices |>
  group_by(symbol) |>
  arrange(date) |>
  mutate(
    retClCl = adjusted / lag(adjusted) - 1,           # Close-to-close (hold overnight)
    retOpCl = (close - open) / close,                 # Open-to-close same day (intraday)
    retClOp = open / lag(adjusted) - 1                # Prior close to open (gap return)
  ) |>
  ungroup() |>
  filter(!is.na(retClCl))

cat("Daily returns computed.\n")
cat("Total rows:", nrow(returns_daily), "\n")
```

## 3.2 Calendar Alignment

```{r alignment}
# Align all series to the SPY trading calendar (removes non-trading days).
spy_calendar <- returns_daily |>
  filter(symbol == "SPY") |>
  pull(date)

returns_daily <- returns_daily |>
  filter(date %in% spy_calendar)

spy_rets_wide <- returns_daily |>
  filter(symbol == "SPY") |>
  select(date, SPY = retClCl)

cat("Calendar aligned to SPY. Trading days:", n_distinct(returns_daily$date), "\n")
```

## 3.3 Train / Test Split

```{r train-test-split}
# Strict temporal split — no testing period data used during model design.
returns_train <- returns_daily |> filter(date >= TRAIN_START, date <= TRAIN_END)
returns_test  <- returns_daily |> filter(date >= TEST_START,  date <= TEST_END)

# Leakage guard
overlap <- intersect(returns_train$date, returns_test$date)
stopifnot("LEAKAGE DETECTED: train and test dates overlap!" = length(overlap) == 0)

cat("Training:", format(TRAIN_START), "to", format(TRAIN_END),
    "—", n_distinct(returns_train$date), "trading days\n")
cat("Testing :", format(TEST_START),  "to", format(TEST_END),
    "—", n_distinct(returns_test$date), "trading days\n")
cat("LEAKAGE CHECK PASS: No date overlap.\n")
```

---

# 4. Feature Engineering

Rolling features capture each ETF's behavioral fingerprint over a 60-day lookback. All features are computed **strictly causally** — only data up to and including each date is used.

```{r feature-engineering}
# ── Rolling helper functions ───────────────────────────────────────────────────

roll_cor <- function(x, y, n) {
  slider::slide2_dbl(x, y,
    .f = ~ cor(.x, .y, use = "complete.obs"),
    .before = n - 1L, .complete = TRUE)
}

roll_beta <- function(x, y, n) {
  slider::slide2_dbl(x, y,
    .f = ~ {
      df <- data.frame(x = .x, y = .y)
      df <- df[complete.cases(df), ]
      if (nrow(df) < 10L) return(NA_real_)
      coef(lm(x ~ y, data = df))[2L]
    },
    .before = n - 1L, .complete = TRUE)
}

roll_cum_ret <- function(x, n) {
  slider::slide_dbl(x,
    .f = ~ prod(1 + .x, na.rm = FALSE) - 1,
    .before = n - 1L, .complete = TRUE)
}

# ── Compute features for all sector ETFs ──────────────────────────────────────
features_daily <- returns_daily |>
  filter(symbol %in% SECTOR_TICKERS) |>
  left_join(spy_rets_wide, by = "date") |>
  group_by(symbol) |>
  arrange(date) |>
  mutate(
    vol_20      = slider::slide_dbl(retClCl, sd, .before = VOL_SHORT - 1L, .complete = TRUE) * sqrt(252),
    vol_60      = slider::slide_dbl(retClCl, sd, .before = VOL_LONG  - 1L, .complete = TRUE) * sqrt(252),
    corr_spy_60 = roll_cor(retClCl, SPY, CORR_WINDOW),
    beta_spy_60 = roll_beta(retClCl, SPY, CORR_WINDOW),
    cum_ret_60  = roll_cum_ret(retClCl, RET_WINDOW),
    spy_cum_60  = roll_cum_ret(SPY, RET_WINDOW),
    rel_ret_60  = cum_ret_60 - spy_cum_60,
    # Also compute Sharpe (for clustering summary, matching class notes style)
    sharpe_60   = slider::slide_dbl(retClCl,
                    ~ if (sd(.x) == 0) NA_real_ else mean(.x) / sd(.x) * sqrt(252),
                    .before = VOL_LONG - 1L, .complete = TRUE)
  ) |>
  ungroup() |>
  select(-SPY, -spy_cum_60)

cat("Feature engineering complete.\n")
```

## 4.1 Feature Completeness

```{r feature-check}
features_daily |>
  filter(date >= TRAIN_START) |>
  group_by(symbol) |>
  summarise(
    pct_vol20    = round(mean(!is.na(vol_20)) * 100, 1),
    pct_corr60   = round(mean(!is.na(corr_spy_60)) * 100, 1),
    pct_cumret60 = round(mean(!is.na(cum_ret_60)) * 100, 1),
    .groups = "drop"
  ) |>
  gt() |>
  tab_header(title = "Feature Completeness in Training Period (%)") |>
  cols_label(symbol = "Ticker", pct_vol20 = "vol_20 %",
             pct_corr60 = "corr_spy_60 %", pct_cumret60 = "cum_ret_60 %")
```

## 4.2 Feature Distributions

```{r feature-dist}
features_daily |>
  filter(date >= TRAIN_START, date <= TRAIN_END) |>
  select(symbol, vol_20, vol_60, corr_spy_60, beta_spy_60, rel_ret_60) |>
  pivot_longer(-symbol, names_to = "feature", values_to = "value") |>
  drop_na() |>
  ggplot(aes(x = value, fill = feature)) +
  geom_histogram(bins = 60, show.legend = FALSE, color = "white", linewidth = 0.1) +
  facet_wrap(~ feature, scales = "free", ncol = 3) +
  scale_fill_viridis_d() +
  labs(title = "Feature Distributions — Training Period (2000–2024)",
       x = "Value", y = "Count") +
  theme_minimal(base_size = 11)
```

---

# 5. Month-End Panel Creation

```{r month-ends}
# Identify the last SPY trading day in each calendar month.
all_trading_days <- sort(unique(returns_daily$date))

month_ends_all <- tibble(date = all_trading_days) |>
  mutate(ym = floor_date(date, "month")) |>
  group_by(ym) |>
  filter(date == max(date)) |>
  ungroup() |>
  pull(date)

train_month_ends <- month_ends_all[month_ends_all >= TRAIN_START & month_ends_all <= TRAIN_END]
test_month_ends  <- month_ends_all[month_ends_all >= TEST_START  & month_ends_all <= TEST_END]
all_month_ends   <- c(train_month_ends, test_month_ends)

cat("Training rebalance dates:", length(train_month_ends), "\n")
cat("Testing  rebalance dates:", length(test_month_ends), "\n")
```

---

# 6. Clustering

Clustering is the **primary signal engine**. At each month-end, sector ETFs are grouped into behaviorally similar clusters using rolling feature snapshots. Cluster membership defines each ETF's comparison set — fair value is cluster-relative, not global.

Following the class notes: we run **both K-Means and Hierarchical** clustering and check for agreement. Agreement = robust structure. Disagreement = investigate further.

## 6.1 Elbow Method: Optimal Number of Clusters

Following the class notes — don't guess k, use the WSS elbow method:

```{r elbow-method}
# Compute within-cluster sum of squares for k = 1 to 6 on a representative snapshot
# Use the most recent training month-end for illustration
last_train_date <- max(train_month_ends)

snap_elbow <- features_daily |>
  filter(date == last_train_date) |>
  select(symbol, all_of(CLUSTER_FEATURES)) |>
  drop_na()

feat_elbow <- as.data.frame(select(snap_elbow, -symbol))
rownames(feat_elbow) <- snap_elbow$symbol

# Standardize (matching recipes::recipe() approach)
scaled_elbow <- scale(feat_elbow)

max_k <- min(6L, nrow(scaled_elbow) - 1L)

wss <- purrr::map_dbl(1:max_k, function(k) {
  stats::kmeans(scaled_elbow, centers = k, nstart = 25L)$tot.withinss
})

tibble(k = 1:max_k, wss = wss) |>
  ggplot(aes(x = k, y = wss)) +
  geom_line(color = "#1565C0", linewidth = 1) +
  geom_point(size = 3, color = "#1565C0") +
  geom_vline(xintercept = N_CLUSTERS, linetype = "dashed", color = "#E53935") +
  scale_x_continuous(breaks = 1:max_k) +
  labs(
    title    = "Elbow Method — Optimal Number of Clusters",
    subtitle = paste0("Red dashed line = chosen k = ", N_CLUSTERS,
                      " | Look for the elbow where improvement slows"),
    x = "Number of Clusters (k)", y = "Within-Cluster Sum of Squares (WSS)"
  ) +
  theme_minimal(base_size = 11)
```

## 6.2 Clustering Functions

```{r clustering-functions}
# ── Standardize features using recipes pipeline (tidymodels preprocessing) ────
standardize_features <- function(df) {
  rec <- recipe(~ ., data = df) |>
    step_normalize(all_numeric_predictors()) |>
    prep(training = df, verbose = FALSE)
  bake(rec, new_data = df)
}

# ── Hierarchical clustering (Ward.D2) ─────────────────────────────────────────
cluster_hierarchical <- function(feature_mat, k = N_CLUSTERS, method = LINK_METHOD) {
  scaled   <- standardize_features(feature_mat)
  dist_mat <- dist(scaled, method = "euclidean")
  hc       <- hclust(dist_mat, method = method)
  list(assignments = cutree(hc, k = k), hclust_obj = hc, dist_mat = dist_mat)
}

# ── K-Means clustering ────────────────────────────────────────────────────────
cluster_kmeans <- function(feature_mat, k = N_CLUSTERS, seed = 42L) {
  scaled <- standardize_features(feature_mat)
  set.seed(seed)
  km <- stats::kmeans(scaled, centers = k, nstart = 25L)
  list(assignments = km$cluster, kmeans_obj = km, scaled = scaled)
}

# ── Apply clustering at a single rebalance date ────────────────────────────────
cluster_at_date <- function(d, features_df, k = N_CLUSTERS) {
  snap <- features_df |>
    filter(date == d) |>
    select(symbol, all_of(CLUSTER_FEATURES)) |>
    drop_na()

  if (nrow(snap) < k) {
    message("Date ", d, ": only ", nrow(snap), " ETFs — skipping.")
    return(NULL)
  }

  feat_mat <- as.data.frame(select(snap, -symbol))
  rownames(feat_mat) <- snap$symbol

  hc_res <- cluster_hierarchical(feat_mat, k = k)
  km_res <- cluster_kmeans(feat_mat, k = k)

  tibble(
    date           = d,
    symbol         = names(hc_res$assignments),
    cluster        = as.integer(hc_res$assignments),
    cluster_kmeans = as.integer(km_res$assignments[names(hc_res$assignments)])
  )
}
```

## 6.3 Apply Clustering at Every Rebalance Date

```{r run-clustering}
cat("Running monthly clustering across", length(all_month_ends), "rebalance dates...\n")

cluster_history <- purrr::map_dfr(
  all_month_ends,
  ~ cluster_at_date(.x, features_daily, k = N_CLUSTERS)
)

# Coverage check
cluster_coverage <- cluster_history |>
  group_by(date) |>
  summarise(n_etfs = n(), n_clusters = n_distinct(cluster), .groups = "drop")

cat("Clustering complete.\n")
cat("Total assignment rows       :", nrow(cluster_history), "\n")
cat("Avg ETFs per rebalance date :", round(mean(cluster_coverage$n_etfs), 1), "\n")
```

## 6.4 Silhouette Analysis: Cluster Quality Validation

Following the class notes — silhouette width > 0.5 suggests reasonable clustering; negative values indicate potential misclassification.

```{r silhouette-analysis}
# Compute silhouette for K-Means and Hierarchical on the most recent training snapshot
snap_sil <- features_daily |>
  filter(date == last_train_date) |>
  select(symbol, all_of(CLUSTER_FEATURES)) |>
  drop_na()

feat_sil <- as.data.frame(select(snap_sil, -symbol))
rownames(feat_sil) <- snap_sil$symbol
scaled_sil  <- as.matrix(standardize_features(feat_sil))
dist_sil    <- dist(scaled_sil)

# Hierarchical silhouette
hc_sil  <- hclust(dist_sil, method = LINK_METHOD)
cuts_hc <- cutree(hc_sil, k = N_CLUSTERS)
sil_hc  <- cluster::silhouette(cuts_hc, dist_sil)

# K-Means silhouette
set.seed(42L)
km_sil  <- stats::kmeans(scaled_sil, centers = N_CLUSTERS, nstart = 25L)
sil_km  <- cluster::silhouette(km_sil$cluster, dist_sil)

p_sil_hc <- factoextra::fviz_silhouette(sil_hc) +
  labs(title = paste("Silhouette — Hierarchical (k =", N_CLUSTERS, ")"),
       subtitle = paste("Avg width:", round(mean(sil_hc[, "sil_width"]), 3),
                        "| >0.5 = well-clustered | <0.3 = weak")) +
  theme_minimal(base_size = 10)

p_sil_km <- factoextra::fviz_silhouette(sil_km) +
  labs(title = paste("Silhouette — K-Means (k =", N_CLUSTERS, ")"),
       subtitle = paste("Avg width:", round(mean(sil_km[, "sil_width"]), 3))) +
  theme_minimal(base_size = 10)

p_sil_hc / p_sil_km

# Agreement check — do both methods agree?
agreement_pct <- mean(cuts_hc == km_sil$cluster) * 100
cat("\nCluster method agreement:", round(agreement_pct, 1), "%\n")
cat("Interpretation: Agreement >= 70% suggests robust structure.\n")
```

## 6.5 K-Means vs Hierarchical Comparison

```{r method-comparison}
# Compare cluster assignments at the last training date
cluster_history |>
  filter(date == last_train_date) |>
  select(symbol, cluster_hclust = cluster, cluster_kmeans) |>
  mutate(agree = if_else(cluster_hclust == cluster_kmeans, "✓ Agree", "✗ Differ")) |>
  gt() |>
  tab_header(
    title    = "K-Means vs Hierarchical Cluster Assignments",
    subtitle = paste("As of", format(last_train_date))
  ) |>
  cols_label(
    symbol         = "ETF",
    cluster_hclust = "Hierarchical",
    cluster_kmeans = "K-Means",
    agree          = "Agreement"
  ) |>
  tab_style(
    style     = cell_fill(color = "#E8F5E9"),
    locations = cells_body(rows = agree == "✓ Agree")
  ) |>
  tab_style(
    style     = cell_fill(color = "#FFEBEE"),
    locations = cells_body(rows = agree == "✗ Differ")
  )
```

## 6.6 Cluster Stability Heatmap

```{r cluster-heatmap}
cluster_history |>
  filter(date >= TRAIN_START, date <= TRAIN_END) |>
  mutate(cluster = factor(cluster)) |>
  ggplot(aes(x = date, y = symbol, fill = cluster)) +
  geom_tile(color = "white", linewidth = 0.15) +
  scale_fill_viridis_d(name = "Cluster") +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  labs(
    title    = "Cluster Assignments Over Time — Training Period",
    subtitle = "Each cell = hierarchical cluster label at that monthly rebalance",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

## 6.7 Dendrogram (Most Recent Training Month-End)

```{r dendrogram}
snap_last <- features_daily |>
  filter(date == last_train_date) |>
  select(symbol, all_of(CLUSTER_FEATURES)) |>
  drop_na()

feat_last <- as.data.frame(select(snap_last, -symbol))
rownames(feat_last) <- snap_last$symbol
scaled_last <- standardize_features(feat_last)
hc_last     <- hclust(dist(scaled_last), method = LINK_METHOD)

fviz_dend(
  hc_last,
  k          = N_CLUSTERS,
  cex        = 0.85,
  k_colors   = viridis(N_CLUSTERS),
  rect       = TRUE,
  rect_fill  = TRUE,
  rect_lty   = 2,
  horiz      = TRUE,
  main       = paste("Cluster Dendrogram —", format(last_train_date)),
  sub        = paste0("Ward.D2 linkage | k = ", N_CLUSTERS, " clusters"),
  xlab       = "Height", ylab = ""
)
```

## 6.8 Cluster Summary Table

Following the class notes — summarise cluster characteristics with color-coded gt table:

```{r cluster-summary-table}
# Compute summary stats per cluster at the most recent training date
snap_summary <- features_daily |>
  filter(date == last_train_date) |>
  select(symbol, vol_20, vol_60, corr_spy_60, beta_spy_60, rel_ret_60, sharpe_60) |>
  drop_na() |>
  left_join(
    cluster_history |> filter(date == last_train_date) |> select(symbol, cluster),
    by = "symbol"
  ) |>
  filter(!is.na(cluster))

snap_summary |>
  group_by(cluster) |>
  summarise(
    n_sectors   = n(),
    avg_vol20   = mean(vol_20),
    avg_corr    = mean(corr_spy_60),
    avg_beta    = mean(beta_spy_60),
    avg_sharpe  = mean(sharpe_60, na.rm = TRUE),
    sectors     = paste(symbol, collapse = ", "),
    .groups     = "drop"
  ) |>
  gt() |>
  tab_header(
    title    = "Cluster Characteristics Summary",
    subtitle = paste("Hierarchical, Ward.D2 | k =", N_CLUSTERS, "| as of", format(last_train_date))
  ) |>
  data_color(
    columns = c(avg_vol20, avg_corr, avg_beta, avg_sharpe),
    palette = c("red", "yellow", "green")
  ) |>
  fmt_percent(columns = avg_vol20, decimals = 1) |>
  fmt_number(columns = c(avg_corr, avg_beta, avg_sharpe), decimals = 2) |>
  fmt_integer(columns = n_sectors) |>
  cols_label(
    cluster    = "Cluster",
    n_sectors  = "# ETFs",
    avg_vol20  = "Ann. Vol (20d)",
    avg_corr   = "Corr to SPY",
    avg_beta   = "Beta to SPY",
    avg_sharpe = "Rolling Sharpe",
    sectors    = "Members"
  )
```

## 6.9 Cluster Stability: Pre vs Post 2020

Following class notes — compare cluster structure across market regimes:

```{r cluster-stability}
# Pre-2020 (excluding COVID) vs Post-2020 feature profiles
pre_2020_date  <- month_ends_all[month_ends_all >= "2019-01-01" &
                                  month_ends_all <= "2019-12-31"] |> max()
post_2020_date <- month_ends_all[month_ends_all >= "2021-01-01" &
                                  month_ends_all <= "2021-12-31"] |> max()

make_regime_snap <- function(d, label) {
  features_daily |>
    filter(date == d) |>
    select(symbol, vol_20, corr_spy_60) |>
    drop_na() |>
    mutate(period = label, date = d)
}

bind_rows(
  make_regime_snap(pre_2020_date,  "Pre-COVID (2019)"),
  make_regime_snap(post_2020_date, "Post-COVID (2021)")
) |>
  ggplot(aes(x = corr_spy_60, y = vol_20, color = period)) +
  geom_point(size = 3.5, alpha = 0.8) +
  geom_text(aes(label = symbol), vjust = -0.8, hjust = 0.5, size = 3) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_color_manual(values = c("Pre-COVID (2019)" = "#1565C0",
                                "Post-COVID (2021)" = "#E65100")) +
  facet_wrap(~ period) +
  labs(
    title    = "Cluster Stability Across Market Regimes",
    subtitle = "Volatility vs SPY correlation — do sector groupings persist?",
    x = "Rolling Correlation to SPY", y = "Ann. Volatility (20d)", color = NULL
  ) +
  theme_minimal(base_size = 11)
```

---

# 7. Cluster-Relative Residual Construction

The residual measures how far an ETF has deviated from its cluster peers over 60 trading days. It is the raw mean-reversion signal before Kalman smoothing.

$$\text{Residual}_{i,t} = \text{CumRet}_{60,i,t} - \underbrace{\frac{1}{|C_i| - 1} \sum_{j \in C_i, j \neq i} \text{CumRet}_{60,j,t}}_{\text{leave-one-out cluster mean}}$$

```{r residual-construction}
# ── Step 1: Carry cluster assignments forward to all daily dates ───────────────
symbol_dates <- features_daily |>
  filter(date >= TRAIN_START) |>
  distinct(date, symbol)

cluster_daily <- symbol_dates |>
  left_join(cluster_history |> select(date, symbol, cluster), by = c("date", "symbol")) |>
  arrange(symbol, date) |>
  group_by(symbol) |>
  fill(cluster, .direction = "down") |>
  ungroup()

# ── Step 2: Join cluster assignments to feature data ──────────────────────────
residuals_daily <- features_daily |>
  filter(date >= TRAIN_START) |>
  select(symbol, date, cum_ret_60) |>
  left_join(cluster_daily, by = c("symbol", "date")) |>
  filter(!is.na(cluster), !is.na(cum_ret_60))

# ── Step 3: Leave-one-out cluster mean ───────────────────────────────────────
residuals_daily <- residuals_daily |>
  group_by(date, cluster) |>
  mutate(
    cluster_sum = sum(cum_ret_60, na.rm = TRUE),
    cluster_n   = sum(!is.na(cum_ret_60)),
    peer_mean   = (cluster_sum - cum_ret_60) / (cluster_n - 1L)
  ) |>
  ungroup() |>
  mutate(residual = cum_ret_60 - peer_mean)

solo_count <- residuals_daily |> filter(cluster_n == 1L) |> nrow()
cat("Solo cluster rows (peer_mean = NA):", solo_count, "\n")
cat("Valid residual rows:", sum(!is.na(residuals_daily$residual)), "\n")
```

## 7.1 Residual Summary

```{r residual-summary}
residuals_daily |>
  filter(!is.na(residual), date >= TRAIN_START, date <= TRAIN_END) |>
  group_by(symbol) |>
  summarise(
    mean_resid = round(mean(residual), 4),
    sd_resid   = round(sd(residual), 4),
    min_resid  = round(min(residual), 4),
    max_resid  = round(max(residual), 4),
    .groups    = "drop"
  ) |>
  gt() |>
  tab_header(title = "Cluster-Relative Residual Summary — Training Period") |>
  cols_label(symbol = "ETF", mean_resid = "Mean", sd_resid = "Std Dev",
             min_resid = "Min", max_resid = "Max")
```

## 7.2 Residual Time Series

```{r residual-viz}
residuals_daily |>
  filter(date >= TRAIN_START, date <= TRAIN_END, !is.na(residual)) |>
  ggplot(aes(x = date, y = residual, color = symbol)) +
  geom_line(alpha = 0.7, linewidth = 0.35) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", linewidth = 0.4) +
  facet_wrap(~ symbol, ncol = 3, scales = "free_y") +
  scale_color_viridis_d(guide = "none") +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  labs(
    title    = "Cluster-Relative Residual Series — Training Period",
    subtitle = "Residual = 60-day compounded return minus leave-one-out cluster peer mean",
    x = NULL, y = "Residual"
  ) +
  theme_minimal(base_size = 10) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

---

# 8. Kalman Filter

The Kalman filter estimates a **smoothed latent fair-value state** from the noisy residual series. The model is a **local level model** (random walk plus noise):

$$\alpha_t = \alpha_{t-1} + \eta_t, \quad \eta_t \sim \mathcal{N}(0, Q)$$
$$y_t = \alpha_t + \varepsilon_t, \quad \varepsilon_t \sim \mathcal{N}(0, R)$$

The **standardized innovation** $z_t = (y_t - \hat{\alpha}_{t|t-1}) / \sqrt{F_t}$ is the trade signal, measuring the model's "surprise" relative to its uncertainty. Per the class notes on mean-reversion and Kalman filter applications, the filter naturally identifies when deviations are large relative to the expected noise level.

## 8.1 Kalman Filter Implementation

```{r kalman-function}
#' Local-level Kalman filter (sequential, causal)
#'
#' @param y  Numeric vector: observed residual series. NAs propagated.
#' @param Q  Process noise variance (state drift speed)
#' @param R  Observation noise variance (smoothing strength)
#' @return List: state, pred_state, pred_var, innovation, z_score
kalman_local_level <- function(y, Q = KF_Q, R = KF_R) {
  n        <- length(y)
  state    <- numeric(n)
  P_filt   <- numeric(n)
  pred_st  <- numeric(n)
  pred_var <- numeric(n)
  innov    <- numeric(n)
  z_score  <- numeric(n)

  first_obs <- which(!is.na(y))[1L]
  if (is.na(first_obs)) {
    return(list(state = rep(NA_real_, n), pred_state = rep(NA_real_, n),
                pred_var = rep(NA_real_, n), innovation = rep(NA_real_, n),
                z_score = rep(NA_real_, n)))
  }

  if (first_obs > 1L) {
    state[seq_len(first_obs - 1L)]    <- NA_real_
    P_filt[seq_len(first_obs - 1L)]   <- NA_real_
    pred_st[seq_len(first_obs - 1L)]  <- NA_real_
    pred_var[seq_len(first_obs - 1L)] <- NA_real_
    innov[seq_len(first_obs - 1L)]    <- NA_real_
    z_score[seq_len(first_obs - 1L)]  <- NA_real_
  }

  state[first_obs]    <- y[first_obs]
  P_filt[first_obs]   <- R
  pred_st[first_obs]  <- y[first_obs]
  pred_var[first_obs] <- R + Q
  innov[first_obs]    <- 0
  z_score[first_obs]  <- 0

  for (t in seq(first_obs + 1L, n)) {
    a_pred <- state[t - 1L]
    P_pred <- P_filt[t - 1L] + Q

    if (is.na(y[t])) {
      state[t]    <- a_pred
      P_filt[t]   <- P_pred
      pred_st[t]  <- a_pred
      pred_var[t] <- P_pred + R
      innov[t]    <- NA_real_
      z_score[t]  <- NA_real_
    } else {
      F_t         <- P_pred + R
      v_t         <- y[t] - a_pred
      K_t         <- P_pred / F_t

      state[t]    <- a_pred + K_t * v_t
      P_filt[t]   <- (1 - K_t) * P_pred
      pred_st[t]  <- a_pred
      pred_var[t] <- F_t
      innov[t]    <- v_t
      z_score[t]  <- v_t / sqrt(F_t)
    }
  }

  list(state = state, pred_state = pred_st, pred_var = pred_var,
       innovation = innov, z_score = z_score)
}
```

## 8.2 Apply Kalman Filter to All ETF Residuals

```{r apply-kalman}
# Filter runs sequentially over full sample (train + test) — strictly causal.
cat("Applying Kalman filter to", n_distinct(residuals_daily$symbol), "ETF residuals...\n")

kalman_results <- residuals_daily |>
  filter(!is.na(residual)) |>
  arrange(symbol, date) |>
  group_by(symbol) |>
  group_modify(~ {
    kf <- kalman_local_level(.x$residual, Q = KF_Q, R = KF_R)
    mutate(.x, kf_state = kf$state, kf_pred_st = kf$pred_state,
           kf_pred_var = kf$pred_var, kf_innov = kf$innovation, kf_z = kf$z_score)
  }) |>
  ungroup()

cat("Kalman filter applied. Valid z-scores:", sum(!is.na(kalman_results$kf_z)), "\n")
```

## 8.3 Kalman Filter Diagnostics

```{r kalman-viz}
# Observed vs filtered state for selected ETFs
selected_etfs <- c("XLE", "XLK", "XLF", "XLU")

kalman_results |>
  filter(symbol %in% selected_etfs, date >= TRAIN_START, date <= TRAIN_END) |>
  pivot_longer(c(residual, kf_state, kf_pred_st),
               names_to = "series", values_to = "value") |>
  mutate(series = recode(series, residual = "Observed Residual",
                         kf_state = "Kalman Filtered State",
                         kf_pred_st = "Predicted State")) |>
  ggplot(aes(x = date, y = value, color = series, alpha = series)) +
  geom_line(linewidth = 0.45) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.3) +
  scale_color_manual(values = c("grey55", "#1565C0", "#E53935")) +
  scale_alpha_manual(values = c(0.4, 0.95, 0.7), guide = "none") +
  facet_wrap(~ symbol, ncol = 2, scales = "free_y") +
  labs(title = "Kalman Filter: Observed Residual vs Smoothed Fair-Value State",
       subtitle = "Blue = filtered state | Grey = observed | Red = predicted",
       x = NULL, y = "Residual", color = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

# Innovation distribution — should approximate N(0,1) if well-specified
kalman_results |>
  filter(date >= TRAIN_START, date <= TRAIN_END, !is.na(kf_z)) |>
  ggplot(aes(x = kf_z)) +
  geom_histogram(aes(y = after_stat(density)), bins = 70,
                 fill = "#1565C0", alpha = 0.7, color = "white", linewidth = 0.1) +
  stat_function(fun = dnorm, color = "#E53935", linewidth = 1) +
  geom_vline(xintercept = c(-SIGNAL_THRESHOLD, SIGNAL_THRESHOLD),
             linetype = "dashed", color = "#FF8F00", linewidth = 0.9) +
  labs(title = "Distribution of Kalman Standardised Innovations — Training Period",
       subtitle = paste0("Orange dashed = ±", SIGNAL_THRESHOLD,
                         " threshold | Red = N(0,1)"),
       x = "z-score", y = "Density") +
  theme_minimal(base_size = 11)
```

---

# 9. Signal Generation

The Kalman z-score at each month-end is the trade signal. Decision rules are explicit and applied uniformly.

| Condition | Signal | Interpretation |
|---|---|---|
| $z < -\tau$ | **LONG** | ETF below Kalman fair value — expect mean reversion upward |
| $z > +\tau$ | **SHORT** | ETF above Kalman fair value — expect mean reversion downward |
| $|z| < \tau$ | **NEUTRAL** | Deviation too small; margin of safety not met |

```{r signal-generation}
signals_monthly <- kalman_results |>
  filter(date %in% all_month_ends, !is.na(kf_z)) |>
  select(date, symbol, residual, kf_state, kf_z, cluster) |>
  arrange(date, desc(abs(kf_z))) |>
  mutate(signal = case_when(
    kf_z < -SIGNAL_THRESHOLD ~ "LONG",
    kf_z >  SIGNAL_THRESHOLD ~ "SHORT",
    TRUE                     ~ "NEUTRAL"
  ))

# Signal distribution (training)
signals_monthly |>
  filter(date >= TRAIN_START, date <= TRAIN_END) |>
  count(signal) |>
  mutate(pct = n / sum(n) * 100) |>
  gt() |>
  tab_header(title = "Signal Distribution — Training Period") |>
  fmt_number(columns = pct, decimals = 1) |>
  cols_label(signal = "Signal", n = "Count", pct = "% of All")

# Signal activity over time
signals_monthly |>
  filter(date >= TRAIN_START, date <= TRAIN_END) |>
  count(date, signal) |>
  ggplot(aes(x = date, y = n, fill = signal)) +
  geom_col(position = "stack", width = 25) +
  scale_fill_manual(values = c(LONG = "#4CAF50", NEUTRAL = "#9E9E9E", SHORT = "#E53935")) +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  labs(title = "Signal Count per Rebalance Date — Training Period",
       subtitle = paste("Threshold |z| ≥", SIGNAL_THRESHOLD),
       x = NULL, y = "Number of ETFs", fill = "Signal") +
  theme_minimal(base_size = 11)
```

---

# 10. Portfolio Construction

```{r portfolio-construction}
# ── Position builder ──────────────────────────────────────────────────────────
build_positions <- function(signals_df,
                            max_longs  = MAX_LONGS,
                            max_shorts = MAX_SHORTS) {
  longs  <- signals_df |> filter(signal == "LONG")  |> arrange(kf_z)        |> head(max_longs)
  shorts <- signals_df |> filter(signal == "SHORT") |> arrange(desc(kf_z))  |> head(max_shorts)
  n_long <- nrow(longs); n_short <- nrow(shorts)

  bind_rows(
    if (n_long  > 0) mutate(longs,  weight =  1 / n_long)  else tibble(),
    if (n_short > 0) mutate(shorts, weight = -1 / n_short) else tibble()
  ) |>
    select(date, symbol, signal, kf_z, cluster, weight)
}

positions_all <- signals_monthly |>
  group_by(date) |>
  group_modify(~ build_positions(.x)) |>
  ungroup()

# Portfolio exposure summary
pos_summary <- positions_all |>
  group_by(date) |>
  summarise(n_long  = sum(signal == "LONG"),
            n_short = sum(signal == "SHORT"),
            gross   = sum(abs(weight)),
            net     = sum(weight), .groups = "drop")

cat("Avg longs per month  :", round(mean(pos_summary$n_long), 1), "\n")
cat("Avg shorts per month :", round(mean(pos_summary$n_short), 1), "\n")
cat("Months with no positions :", sum(pos_summary$gross == 0), "\n")

# Gross/net exposure chart
pos_summary |>
  filter(date >= TRAIN_START) |>
  pivot_longer(c(gross, net), names_to = "type", values_to = "exposure") |>
  ggplot(aes(x = date, y = exposure, color = type)) +
  geom_line(linewidth = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = TEST_START, linetype = "dotted", color = "black") +
  scale_color_manual(values = c(gross = "#1565C0", net = "#E53935")) +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  labs(title = "Portfolio Gross and Net Exposure per Rebalance",
       subtitle = "Dotted = start of testing period",
       x = NULL, y = "Exposure", color = "Type") +
  theme_minimal(base_size = 11)
```

## 10.1 Position Chart — ETF Weights Over Time

```{r position-chart}
# Build a full grid: every sector ETF × every rebalance date (post TRAIN_START)
all_symbols_pos <- expand_grid(
  symbol = SECTOR_TICKERS,
  date   = all_month_ends[all_month_ends >= TRAIN_START]
)

positions_full_grid <- all_symbols_pos |>
  left_join(
    positions_all |> select(date, symbol, weight, signal),
    by = c("date", "symbol")
  ) |>
  mutate(weight = replace_na(weight, 0))

# Heatmap: green = long, red = short, white = flat
ggplot(positions_full_grid, aes(x = date, y = symbol, fill = weight)) +
  geom_tile(color = "white", linewidth = 0.1) +
  scale_fill_gradient2(
    low      = "#E53935",
    mid      = "white",
    high     = "#4CAF50",
    midpoint = 0,
    name     = "Weight",
    labels   = percent_format(accuracy = 1)
  ) +
  geom_vline(xintercept = as.numeric(TEST_START),
             linetype = "dotted", linewidth = 0.7, color = "black") +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  labs(
    title    = "ETF Position Weights Over Time",
    subtitle = "Green = long | Red = short | White = flat | Dotted line = start of testing period",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

---

# 11. Backtest

## 11.1 Monthly Returns

```{r monthly-returns}
monthly_prices <- prices |>
  filter(date %in% month_ends_all) |>
  arrange(symbol, date)

monthly_ret_tbl <- monthly_prices |>
  group_by(symbol) |>
  arrange(date) |>
  mutate(monthly_ret = adjusted / lag(adjusted) - 1) |>
  ungroup() |>
  filter(!is.na(monthly_ret))

cat("Monthly returns computed for", n_distinct(monthly_ret_tbl$symbol), "symbols.\n")
```

## 11.2 Strategy Return Calculation

```{r backtest}
# Signal at month-end d → position held in following month d+1.
# Return(d+1) = strategy weighted ETF returns from close d to close d+1.
next_month_map <- tibble(
  signal_date = all_month_ends[-length(all_month_ends)],
  return_date = all_month_ends[-1]
)

strategy_returns_raw <- positions_all |>
  rename(signal_date = date) |>
  left_join(next_month_map, by = "signal_date") |>
  filter(!is.na(return_date)) |>
  left_join(monthly_ret_tbl |> select(symbol, return_date = date, monthly_ret),
            by = c("symbol", "return_date")) |>
  filter(!is.na(monthly_ret)) |>
  group_by(return_date) |>
  summarise(strategy_ret = sum(weight * monthly_ret, na.rm = TRUE),
            n_positions  = n(), .groups = "drop")

all_return_months <- tibble(return_date = all_month_ends[-1])
strategy_returns <- all_return_months |>
  left_join(strategy_returns_raw, by = "return_date") |>
  mutate(strategy_ret = replace_na(strategy_ret, 0),
         n_positions  = replace_na(n_positions, 0)) |>
  filter(return_date >= TRAIN_START)

spy_monthly <- monthly_ret_tbl |>
  filter(symbol == "SPY") |>
  select(return_date = date, spy_ret = monthly_ret)

backtest_df <- strategy_returns |>
  left_join(spy_monthly, by = "return_date") |>
  filter(!is.na(spy_ret)) |>
  mutate(period = if_else(return_date <= TRAIN_END, "Training", "Testing"))

cat("Training months:", sum(backtest_df$period == "Training"), "\n")
cat("Testing  months:", sum(backtest_df$period == "Testing"), "\n")
```

## 11.3 Strategy Chart (Price, Positions, Cumulative Equity)

Following the class notes — convert to xts and use multi-panel chart:

```{r strategy-chart}
# Build monthly xts of strategy and SPY returns
backtest_xts <- backtest_df |>
  select(return_date, strategy_ret, spy_ret) |>
  mutate(cum_strat = cumprod(1 + strategy_ret),
         cum_spy   = cumprod(1 + spy_ret)) |>
  timetk::tk_xts(date_var = return_date)

# Plot cumulative equity (course-style xts chart)
plot(backtest_xts$cum_strat,
     main  = "Cumulative Equity — Strategy vs SPY",
     col   = "#1565C0",
     ylab  = "Growth of $1")
xts::addSeries(backtest_xts$cum_spy,
     main = "SPY", on = NA,
     type = "l", col = "#E65100", lty = 2, lwd = 1.5, pch = 0)
xts::addSeries(backtest_xts$strategy_ret,
     main = "Monthly Return", on = NA,
     type = "h", col = "#1565C0", lty = 1, lwd = 1, pch = 0)
```

## 11.4 Trade Markers on Price Chart

Following the class notes style — overlay entry/exit markers directly on the price chart to visually confirm signal timing and frequency.

```{r trade-markers}
# SPY monthly price as the backdrop
spy_prices_monthly <- monthly_prices |>
  filter(symbol == "SPY", date >= TRAIN_START) |>
  select(date, adjusted)

# Summarise position direction at each rebalance
position_markers <- positions_all |>
  filter(date >= TRAIN_START) |>
  group_by(date) |>
  summarise(
    n_long  = sum(signal == "LONG"),
    n_short = sum(signal == "SHORT"),
    .groups = "drop"
  ) |>
  mutate(direction = case_when(
    n_long > 0 & n_short > 0  ~ "Long & Short",
    n_long > 0                 ~ "Long Only",
    n_short > 0                ~ "Short Only",
    TRUE                       ~ "Flat"
  )) |>
  filter(direction != "Flat") |>
  left_join(spy_prices_monthly, by = "date")

ggplot(spy_prices_monthly, aes(x = date, y = adjusted)) +
  geom_line(color = "grey55", linewidth = 0.5) +
  # Long entries (upward triangles)
  geom_point(
    data  = position_markers |> filter(grepl("Long", direction)),
    aes(color = "Long Entry"),
    shape = 24, size = 2.8, fill = "#4CAF50", stroke = 0.4
  ) +
  # Short entries (downward triangles)
  geom_point(
    data  = position_markers |> filter(grepl("Short", direction)),
    aes(color = "Short Entry"),
    shape = 25, size = 2.8, fill = "#E53935", stroke = 0.4
  ) +
  geom_vline(xintercept = TEST_START, linetype = "dotted",
             color = "black", linewidth = 0.7) +
  annotate("text", x = TEST_START + 30, y = min(spy_prices_monthly$adjusted, na.rm = TRUE),
           label = "← Train | Test →", size = 3, hjust = 0, color = "grey40") +
  scale_color_manual(
    values = c("Long Entry" = "#4CAF50", "Short Entry" = "#E53935"),
    guide  = guide_legend(
      override.aes = list(shape = c(24L, 25L), fill = c("#4CAF50", "#E53935"))
    )
  ) +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  scale_y_continuous(labels = dollar_format()) +
  labs(
    title    = "Trade Markers on SPY Price Chart",
    subtitle = "▲ = rebalance with ≥1 long | ▼ = rebalance with ≥1 short | Monthly frequency",
    x = NULL, y = "SPY Adjusted Price ($)", color = "Trade Direction"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

---

# 12. Evaluation and Charts

## 12.1 Cumulative Returns

```{r cum-returns}
backtest_df |>
  select(return_date, strategy_ret, spy_ret) |>
  pivot_longer(-return_date, names_to = "series", values_to = "ret") |>
  group_by(series) |>
  mutate(cum_ret = cumprod(1 + ret) - 1) |>
  ungroup() |>
  mutate(series = recode(series,
    strategy_ret = "Strategy — Cluster Mean Reversion",
    spy_ret      = "SPY Benchmark")) |>
  ggplot(aes(x = return_date, y = cum_ret, color = series, linetype = series)) +
  geom_line(linewidth = 0.9) +
  geom_vline(xintercept = TEST_START, linetype = "dotted",
             color = "black", linewidth = 0.7) +
  annotate("text", x = TEST_START + 20, y = -0.05,
           label = "← Training  |  Testing →", size = 3.2, hjust = 0, color = "grey30") +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_color_manual(values = c("#1565C0", "#E65100")) +
  scale_linetype_manual(values = c("solid", "dashed")) +
  labs(title = "Cumulative Returns: Strategy vs SPY Benchmark",
       x = NULL, y = "Cumulative Return", color = NULL, linetype = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

## 12.2 Drawdown Analysis — PerformanceAnalytics

Following the class notes — use PerformanceAnalytics for drawdown charts and top drawdown tables:

```{r drawdown-pa}
# Convert strategy and SPY returns to xts for PerformanceAnalytics
strat_xts <- backtest_df |>
  select(return_date, strategy_ret) |>
  timetk::tk_xts(date_var = return_date)

spy_xts <- backtest_df |>
  select(return_date, spy_ret) |>
  timetk::tk_xts(date_var = return_date)

# PerformanceAnalytics drawdown chart
PerformanceAnalytics::chart.Drawdown(
  cbind(strat_xts, spy_xts),
  legend.loc = "bottomleft",
  main       = "Drawdown: Strategy vs SPY",
  colorset   = c("#1565C0", "#E65100"),
  ylab       = "Drawdown"
)

# Top drawdowns table (class notes style: table.Drawdowns)
cat("\n--- Top 10 Strategy Drawdowns ---\n")
PerformanceAnalytics::table.Drawdowns(strat_xts, top = 10)
```

## 12.3 Correlation to Underlying (SPY)

From the class notes: when ρ = 1, strategy equals long buy-and-hold; ρ = -1 equals short. Values between reveal the strategy's independence from SPY.

```{r correlation-chart}
strat_vs_spy <- cbind(strat_xts, spy_xts)
colnames(strat_vs_spy) <- c("Strategy", "SPY")

PerformanceAnalytics::chart.Correlation(
  strat_vs_spy,
  histogram = TRUE,
  main      = "Strategy vs SPY — Return Correlation"
)
```

## 12.4 tradeStats — Full Performance Summary

Mirroring the `RTL::tradeStats()` format from the class notes: CumReturn, Ret.Ann, SD.Ann, Sharpe, Omega, %.Win, %.InMrkt, DD.Max, DD.Length.

```{r trade-stats-function}
# Custom tradeStats function matching RTL::tradeStats() output format
trade_stats <- function(ret_vec, rf = 0) {
  r       <- ret_vec[!is.na(ret_vec)]
  n       <- length(r)
  if (n < 3L) return(tibble())

  cum_ret  <- prod(1 + r) - 1
  ret_ann  <- prod(1 + r)^(12 / n) - 1
  sd_ann   <- sd(r) * sqrt(12)
  sharpe   <- (ret_ann - rf) / sd_ann

  # Omega (MAR = 0, matching class notes)
  MAR       <- 0
  gains     <- r[r > MAR]
  losses    <- r[r <= MAR]
  omega     <- if (length(losses) > 0 && sum(abs(losses)) > 0)
                 sum(gains - MAR) / sum(MAR - losses) else Inf

  pct_win   <- mean(r > 0)
  pct_in_mkt <- mean(r != 0)

  # Max drawdown depth and length
  cum_growth <- cumprod(1 + r)
  running_max <- cummax(cum_growth)
  dd          <- (cum_growth - running_max) / running_max
  dd_max      <- min(dd)

  # Drawdown length: max consecutive periods below peak
  below_peak <- dd < 0
  dd_lengths <- rle(below_peak)
  dd_length  <- if (any(dd_lengths$values)) max(dd_lengths$lengths[dd_lengths$values]) else 0L

  tibble(
    CumReturn  = cum_ret,
    Ret.Ann    = ret_ann,
    SD.Ann     = sd_ann,
    Sharpe     = sharpe,
    Omega      = omega,
    `%.Win`    = pct_win,
    `%.InMrkt` = pct_in_mkt,
    DD.Length  = dd_length,
    DD.Max     = dd_max,
    N.Trades   = sum(r != 0),   # Months with an active position
    N.Months   = n
  )
}
```

```{r trade-stats-output}
# Compute for all period / series combinations
stats_list <- list(
  "Strategy — Training" = filter(backtest_df, period == "Training")$strategy_ret,
  "Strategy — Testing"  = filter(backtest_df, period == "Testing")$strategy_ret,
  "Strategy — Full"     = backtest_df$strategy_ret,
  "SPY — Training"      = filter(backtest_df, period == "Training")$spy_ret,
  "SPY — Testing"       = filter(backtest_df, period == "Testing")$spy_ret,
  "SPY — Full"          = backtest_df$spy_ret
)

stats_tbl <- purrr::imap_dfr(stats_list, ~ mutate(trade_stats(.x), Period = .y, .before = 1))

stats_tbl |>
  gt() |>
  tab_header(
    title    = "Performance Summary — tradeStats",
    subtitle = "Monthly returns | annualised metrics | matching class notes format"
  ) |>
  fmt_percent(columns = c(CumReturn, Ret.Ann, SD.Ann, `%.Win`, `%.InMrkt`, DD.Max),
              decimals = 1) |>
  fmt_number(columns  = c(Sharpe, Omega), decimals = 2) |>
  fmt_integer(columns = c(DD.Length, N.Trades, N.Months)) |>
  tab_style(
    style     = cell_fill(color = "#E3F2FD"),
    locations = cells_body(rows = grepl("Strategy", Period))
  ) |>
  cols_label(
    Period     = "Series",
    CumReturn  = "Cum Return",
    Ret.Ann    = "Ann. Return",
    SD.Ann     = "Ann. Vol",
    Sharpe     = "Sharpe",
    Omega      = "Omega",
    `%.Win`    = "% Win",
    `%.InMrkt` = "% In Mkt",
    DD.Length  = "DD Length",
    DD.Max     = "DD Max",
    N.Trades   = "# Trades",
    N.Months   = "Months"
  )
```

## 12.5 Sharpe and Omega via PerformanceAnalytics

```{r pa-metrics}
# PerformanceAnalytics versions (matching class notes)
cat("=== Strategy - Training Period ===\n")
train_xts <- strat_xts[index(strat_xts) <= TRAIN_END]

cat("Sharpe Ratio:\n")
print(PerformanceAnalytics::SharpeRatio(train_xts))

cat("\nOmega Sharpe Ratio (MAR = 0):\n")
print(PerformanceAnalytics::OmegaSharpeRatio(train_xts, MAR = 0))

cat("\n=== Strategy - Testing Period ===\n")
test_xts <- strat_xts[index(strat_xts) >= TEST_START]
print(PerformanceAnalytics::SharpeRatio(test_xts))
```

## 12.6 Rolling 12-Month Sharpe

```{r rolling-sharpe}
backtest_df <- backtest_df |>
  mutate(
    roll_sharpe_strat = slider::slide_dbl(
      strategy_ret,
      ~ if (sd(.x) == 0) NA_real_ else mean(.x) / sd(.x) * sqrt(12),
      .before = 11L, .complete = TRUE),
    roll_sharpe_spy = slider::slide_dbl(
      spy_ret,
      ~ if (sd(.x) == 0) NA_real_ else mean(.x) / sd(.x) * sqrt(12),
      .before = 11L, .complete = TRUE)
  )

backtest_df |>
  filter(!is.na(roll_sharpe_strat)) |>
  select(return_date, roll_sharpe_strat, roll_sharpe_spy) |>
  pivot_longer(-return_date, names_to = "series", values_to = "sharpe") |>
  mutate(series = recode(series,
    roll_sharpe_strat = "Strategy", roll_sharpe_spy = "SPY")) |>
  ggplot(aes(x = return_date, y = sharpe, color = series)) +
  geom_line(linewidth = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_hline(yintercept = 1, linetype = "dotted", color = "grey50") +
  geom_vline(xintercept = TEST_START, linetype = "dotted", color = "black") +
  scale_color_manual(values = c(Strategy = "#1565C0", SPY = "#E65100")) +
  labs(title = "12-Month Rolling Sharpe Ratio",
       x = NULL, y = "Sharpe (annualised)", color = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

# Rolling 12-month cumulative return
backtest_df |>
  mutate(
    roll_ret_strat = slider::slide_dbl(strategy_ret,
                       ~ prod(1 + .x) - 1, .before = 11L, .complete = TRUE),
    roll_ret_spy   = slider::slide_dbl(spy_ret,
                       ~ prod(1 + .x) - 1, .before = 11L, .complete = TRUE)
  ) |>
  filter(!is.na(roll_ret_strat)) |>
  select(return_date, roll_ret_strat, roll_ret_spy) |>
  pivot_longer(-return_date, names_to = "series", values_to = "ret") |>
  mutate(series = recode(series,
    roll_ret_strat = "Strategy", roll_ret_spy = "SPY")) |>
  ggplot(aes(x = return_date, y = ret, color = series)) +
  geom_line(linewidth = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = TEST_START, linetype = "dotted", color = "black") +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_color_manual(values = c(Strategy = "#1565C0", SPY = "#E65100")) +
  labs(
    title    = "12-Month Rolling Cumulative Return",
    subtitle = "Dotted = start of testing period",
    x = NULL, y = "12-Month Return", color = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

## 12.7 Sharpe vs Omega: Return Distribution Overlay

Following the class notes: Sharpe assumes returns are normally distributed and penalises upside and downside equally. Omega measures the actual ratio of gains to losses — it captures skew and fat tails that Sharpe misses. This chart visualises **why both matter**.

```{r sharpe-omega-overlay}
# Training period returns for the visual
ret_train_vec <- backtest_df |>
  filter(period == "Training") |>
  pull(strategy_ret)

ret_df_viz <- tibble(ret = ret_train_vec) |> drop_na()

mu_train    <- mean(ret_df_viz$ret)
sigma_train <- sd(ret_df_viz$ret)
MAR         <- 0

# Omega numerator / denominator from actual returns
omega_num <- sum(ret_df_viz$ret[ret_df_viz$ret > MAR] - MAR)
omega_den <- sum(MAR - ret_df_viz$ret[ret_df_viz$ret <= MAR])
omega_val <- if (omega_den > 0) omega_num / omega_den else Inf

sharpe_ann <- (mu_train * 12) / (sigma_train * sqrt(12))

ggplot(ret_df_viz, aes(x = ret)) +
  # Gains (Omega numerator shaded green)
  geom_histogram(
    data    = filter(ret_df_viz, ret > MAR),
    aes(y = after_stat(density), fill = "Gains — Omega numerator"),
    bins = 25, alpha = 0.75, color = "white", linewidth = 0.1
  ) +
  # Losses (Omega denominator shaded red)
  geom_histogram(
    data    = filter(ret_df_viz, ret <= MAR),
    aes(y = after_stat(density), fill = "Losses — Omega denominator"),
    bins = 25, alpha = 0.75, color = "white", linewidth = 0.1
  ) +
  # Normal curve (Sharpe's assumed distribution)
  stat_function(
    fun  = dnorm,
    args = list(mean = mu_train, sd = sigma_train),
    color = "#1565C0", linewidth = 1.3, linetype = "solid"
  ) +
  # Mean line (Sharpe numerator)
  geom_vline(xintercept = mu_train, color = "#1565C0",
             linetype = "dashed", linewidth = 1) +
  # MAR = 0
  geom_vline(xintercept = MAR, color = "black", linewidth = 0.6) +
  # ±1 SD shading bands
  geom_vline(xintercept = c(mu_train - sigma_train, mu_train + sigma_train),
             color = "#1565C0", linetype = "dotted", linewidth = 0.7) +
  scale_fill_manual(values = c(
    "Gains — Omega numerator"   = "#4CAF50",
    "Losses — Omega denominator" = "#E53935"
  )) +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  annotate("text", x = mu_train + 0.0015, y = Inf,
           label = paste0("μ  (Sharpe numerator)"),
           vjust = 2.2, hjust = 0, color = "#1565C0", size = 3.2) +
  labs(
    title    = "Sharpe vs Omega: Return Distribution — Training Period",
    subtitle = paste0(
      "Blue curve = Normal(μ, σ) assumed by Sharpe  |  ",
      "Omega = Green area ÷ Red area\n",
      "Sharpe (ann.) = ", round(sharpe_ann, 2),
      "  |  Omega = ", round(omega_val, 2),
      "  |  Dashed = mean  |  Dotted = ±1 SD"
    ),
    x = "Monthly Return", y = "Density", fill = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

## 12.8 Trade Analysis

```{r trade-analysis}
# Turnover
turnover_df <- positions_all |>
  arrange(symbol, date) |>
  group_by(symbol) |>
  mutate(weight_lag = lag(weight)) |>
  ungroup() |>
  group_by(date) |>
  summarise(
    turnover = sum(abs(replace_na(weight, 0) - replace_na(weight_lag, 0))) / 2,
    .groups  = "drop")

cat("Avg monthly one-way turnover:",
    percent(mean(turnover_df$turnover, na.rm = TRUE), accuracy = 0.1), "\n")

# Win/loss by signal direction
position_returns_detail <- positions_all |>
  rename(signal_date = date) |>
  left_join(next_month_map, by = "signal_date") |>
  filter(!is.na(return_date)) |>
  left_join(monthly_ret_tbl |> select(symbol, return_date = date, monthly_ret),
            by = c("symbol", "return_date")) |>
  filter(!is.na(monthly_ret)) |>
  mutate(pnl = weight * monthly_ret, won = pnl > 0)

position_returns_detail |>
  group_by(signal) |>
  summarise(
    n_trades    = n(),
    hit_rate    = mean(won) * 100,
    avg_pnl_pct = mean(pnl) * 100,
    med_pnl_pct = median(pnl) * 100,
    best_pct    = max(pnl) * 100,
    worst_pct   = min(pnl) * 100,
    .groups = "drop"
  ) |>
  gt() |>
  tab_header(title = "Trade Analysis by Signal Direction — Full Sample") |>
  fmt_integer(columns = n_trades) |>
  fmt_number(columns = c(hit_rate, avg_pnl_pct, med_pnl_pct, best_pct, worst_pct),
             decimals = 2) |>
  cols_label(signal = "Signal", n_trades = "# Trades", hit_rate = "Hit Rate %",
             avg_pnl_pct = "Avg P&L %", med_pnl_pct = "Median P&L %",
             best_pct = "Best %", worst_pct = "Worst %")

# Return distribution
backtest_df |>
  ggplot(aes(x = strategy_ret, fill = period)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40, alpha = 0.75,
                 position = "identity", color = "white", linewidth = 0.1) +
  stat_function(fun = dnorm,
                args = list(mean = mean(backtest_df$strategy_ret),
                            sd   = sd(backtest_df$strategy_ret)),
                color = "red", linewidth = 1) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_fill_manual(values = c(Training = "#1565C0", Testing = "#E65100")) +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  labs(title = "Monthly Return Distribution — Strategy (vs Normal)",
       subtitle = "Can you tell why Omega differs from Sharpe here?",
       x = "Monthly Return", y = "Density", fill = "Period") +
  theme_minimal(base_size = 11)
```

## 12.9 Training vs Testing Comparison

```{r train-test-comparison}
backtest_df |>
  select(return_date, period, strategy_ret, spy_ret) |>
  group_by(period) |>
  mutate(cum_strat = cumprod(1 + strategy_ret) - 1,
         cum_spy   = cumprod(1 + spy_ret) - 1) |>
  ungroup() |>
  pivot_longer(c(cum_strat, cum_spy), names_to = "series", values_to = "cum_ret") |>
  mutate(series = recode(series, cum_strat = "Strategy", cum_spy = "SPY")) |>
  ggplot(aes(x = return_date, y = cum_ret, color = series)) +
  geom_line(linewidth = 0.85) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~ period, scales = "free") +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_color_manual(values = c(Strategy = "#1565C0", SPY = "#E65100")) +
  labs(title = "Cumulative Returns by Period (Rebased to Zero)",
       x = NULL, y = "Cumulative Return", color = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

---

# 13. Parameter Sensitivity

Robust strategies show stable performance across parameter ranges. Cliff-like optima suggest overfitting. Following the class notes: we visualize results as a grid and use Z-score normalization across metrics for comparability.

## 13.1 Signal Threshold Sensitivity

```{r sensitivity-threshold}
thresholds <- seq(0.50, 2.00, by = 0.25)

sensitivity_tau <- purrr::map_dfr(thresholds, function(tau) {
  sigs <- kalman_results |>
    filter(date %in% train_month_ends, !is.na(kf_z)) |>
    select(date, symbol, kf_z) |>
    mutate(signal = case_when(
      kf_z < -tau ~ "LONG", kf_z > tau ~ "SHORT", TRUE ~ "NEUTRAL"))

  pos <- sigs |>
    group_by(date) |>
    group_modify(~ build_positions(.x)) |>
    ungroup()

  strat_ret <- pos |>
    rename(signal_date = date) |>
    left_join(next_month_map |> filter(signal_date %in% train_month_ends),
              by = "signal_date") |>
    filter(!is.na(return_date)) |>
    left_join(monthly_ret_tbl |> select(symbol, return_date = date, monthly_ret),
              by = c("symbol", "return_date")) |>
    filter(!is.na(monthly_ret)) |>
    group_by(return_date) |>
    summarise(strategy_ret = sum(weight * monthly_ret, na.rm = TRUE), .groups = "drop")

  if (nrow(strat_ret) < 12L) return(tibble(tau = tau, Sharpe = NA, CumReturn = NA, DD.Max = NA))

  m <- trade_stats(strat_ret$strategy_ret)
  tibble(tau = tau, Sharpe = m$Sharpe, CumReturn = m$CumReturn, DD.Max = m$DD.Max)
})

# Z-score normalization across metrics (class notes style)
sensitivity_Z <- sensitivity_tau |>
  pivot_longer(-tau, names_to = "metric", values_to = "value") |>
  group_by(metric) |>
  mutate(valueZ = (value - mean(value, na.rm = TRUE)) / sd(value, na.rm = TRUE)) |>
  ungroup()

# Heatmap of Z-score normalized metrics (class notes style)
sensitivity_Z |>
  ggplot(aes(x = tau, y = metric, fill = valueZ)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "red", mid = "white", high = "blue", midpoint = 0,
                       name = "Z-score") +
  geom_text(aes(label = round(value, 2)), size = 3) +
  labs(
    title    = "Sensitivity Grid — Signal Threshold (τ) — Training Period",
    subtitle = "Z-score normalized for comparability | Red dashed = chosen threshold",
    x = "Signal Threshold (τ)", y = "Metric"
  ) +
  geom_vline(xintercept = SIGNAL_THRESHOLD, linetype = "dashed", color = "#E53935",
             linewidth = 0.8) +
  theme_minimal(base_size = 11)

# Line chart for clarity
sensitivity_tau |>
  pivot_longer(-tau, names_to = "metric", values_to = "value") |>
  ggplot(aes(x = tau, y = value)) +
  geom_line(color = "#1565C0", linewidth = 1) +
  geom_point(size = 2.5) +
  geom_vline(xintercept = SIGNAL_THRESHOLD, linetype = "dashed", color = "#E53935") +
  facet_wrap(~ metric, scales = "free_y", ncol = 3,
             labeller = labeller(metric = c(
               CumReturn = "Cumulative Return",
               DD.Max    = "Max Drawdown",
               Sharpe    = "Sharpe Ratio"))) +
  labs(title = "Parameter Sensitivity: Signal Threshold (τ) — Training Period",
       subtitle = "Prefer stable region, not cliff-like optima",
       x = "Signal Threshold (τ)", y = "Value") +
  theme_minimal(base_size = 11)
```

## 13.2 Table of Top Parameter Combinations

Following the class notes — rank all parameter combinations and highlight the top performers. Prefer stable regions over cliff-like optima (neighboring values should also score well).

```{r top-parameters}
# Full ranked table of all parameter combinations tested
sensitivity_tau |>
  arrange(desc(Sharpe)) |>
  mutate(
    rank    = row_number(),
    optimal = if_else(rank == 1L, "★ Best", ""),
    chosen  = if_else(tau == SIGNAL_THRESHOLD, "← Chosen", "")
  ) |>
  gt() |>
  tab_header(
    title    = "All Parameter Combinations — Ranked by Sharpe (Training Period)",
    subtitle = paste0(
      "Prefer stable region over cliff-like optima | Chosen τ = ",
      SIGNAL_THRESHOLD, " | ★ = best Sharpe"
    )
  ) |>
  fmt_percent(columns = c(CumReturn, DD.Max), decimals = 1) |>
  fmt_number(columns = Sharpe, decimals = 3) |>
  fmt_integer(columns = rank) |>
  tab_style(
    style     = cell_fill(color = "#FFF9C4"),
    locations = cells_body(rows = chosen == "← Chosen")
  ) |>
  tab_style(
    style     = cell_fill(color = "#E8F5E9"),
    locations = cells_body(rows = optimal == "★ Best")
  ) |>
  cols_label(
    rank      = "Rank",
    tau       = "Threshold (τ)",
    Sharpe    = "Sharpe",
    CumReturn = "Cum Return",
    DD.Max    = "Max DD",
    optimal   = "",
    chosen    = ""
  )
```

## 13.3 Risk Appetite Filter

Following the class notes — filter by explicit risk appetite rules, then rank. Useful when multiple metrics conflict: Sharpe vs drawdown tradeoff.

```{r risk-appetite}
# Define risk appetite: Sharpe > 0 and Max Drawdown > -40%
sensitivity_tau |>
  filter(!is.na(Sharpe), Sharpe > 0, DD.Max > -0.40) |>
  arrange(desc(Sharpe)) |>
  gt() |>
  tab_header(
    title    = "Parameter Combinations Meeting Risk Appetite",
    subtitle = "Filter: Sharpe > 0 and DD.Max > -40% | Ranked by Sharpe"
  ) |>
  fmt_percent(columns = c(CumReturn, DD.Max), decimals = 1) |>
  fmt_number(columns = Sharpe, decimals = 3) |>
  cols_label(tau = "Threshold (τ)", Sharpe = "Sharpe",
             CumReturn = "Cum Return", DD.Max = "Max DD")
```

---

# 14. Limitations and Assumptions

## Transaction Costs and Turnover

Transaction costs are not modeled in the backtest returns above, but are explicitly quantified here as required by the class notes.

```{r transaction-costs}
# ── Turnover summary ────────────────────────────────────────────────────────────
turnover_summary <- turnover_df |>
  filter(date >= TRAIN_START) |>
  summarise(
    avg_monthly_turnover    = mean(turnover, na.rm = TRUE),
    median_monthly_turnover = median(turnover, na.rm = TRUE),
    max_monthly_turnover    = max(turnover, na.rm = TRUE),
    annual_turnover_est     = mean(turnover, na.rm = TRUE) * 12
  )

cat("=== Turnover Analysis ===\n")
cat("Avg monthly one-way turnover  :", percent(turnover_summary$avg_monthly_turnover, 0.1), "\n")
cat("Median monthly one-way turnover:", percent(turnover_summary$median_monthly_turnover, 0.1), "\n")
cat("Est. annual one-way turnover  :", percent(turnover_summary$annual_turnover_est, 0.1), "\n")

# ── Estimate performance drag at different cost assumptions ────────────────────
cost_bps <- c(5, 10, 20, 30)   # Round-trip transaction cost in basis points

cost_impact <- tibble(
  cost_bps     = cost_bps,
  annual_drag  = turnover_summary$annual_turnover_est * (cost_bps / 10000),
  cost_pct     = cost_bps / 100
) |>
  mutate(
    description = paste0(cost_bps, " bps round-trip"),
    annual_drag_pct = annual_drag
  )

cost_impact |>
  select(description, cost_bps, annual_drag_pct) |>
  gt() |>
  tab_header(
    title    = "Estimated Annual Performance Drag from Transaction Costs",
    subtitle = paste0(
      "Based on estimated annual one-way turnover of ",
      percent(turnover_summary$annual_turnover_est, 0.1),
      " | Sector SPDR ETFs: typical bid/ask ~1–3 bps"
    )
  ) |>
  fmt_percent(columns = annual_drag_pct, decimals = 2) |>
  cols_label(
    description    = "Assumed Cost",
    cost_bps       = "Round-Trip (bps)",
    annual_drag_pct = "Est. Annual Drag"
  ) |>
  tab_footnote(
    footnote  = "Drag = annual turnover × round-trip cost. Sector SPDRs are highly liquid; 5–10 bps is realistic.",
    locations = cells_column_labels(columns = annual_drag_pct)
  )

# ── Turnover over time chart ────────────────────────────────────────────────────
turnover_df |>
  filter(date >= TRAIN_START) |>
  ggplot(aes(x = date, y = turnover)) +
  geom_col(fill = "#1565C0", alpha = 0.7, width = 25) +
  geom_hline(yintercept = mean(turnover_df$turnover, na.rm = TRUE),
             linetype = "dashed", color = "#E53935") +
  geom_vline(xintercept = TEST_START, linetype = "dotted", color = "black") +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  annotate("text", x = TEST_START + 30, y = 0,
           label = "← Train | Test →", size = 3, hjust = 0, color = "grey40") +
  labs(
    title    = "Monthly Portfolio Turnover",
    subtitle = "Red dashed = average | Dotted = start of testing period",
    x = NULL, y = "One-Way Turnover"
  ) +
  theme_minimal(base_size = 11)
```

## Implementation Assumptions

| Assumption | Detail |
|---|---|
| **Execution timing** | Signal at month-end close; return from that close to the next month-end close. No intraday timing assumed. |
| **Short selling** | ETFs assumed shortable at zero cost. Borrow fees, dividends owed on shorts, and margin mechanics excluded (permitted by assignment scope per class Q&A). |
| **Transaction costs** | Not modeled. Sector SPDR ETFs are highly liquid — bid/offer and commissions are small but non-zero. |
| **Kalman initialization** | Filter initialized at first non-NA residual. Early estimates less reliable for first ~60 days per series. |
| **Solo clusters** | Singleton clusters (cluster_n = 1) produce undefined peer means and are excluded from signal generation. |
| **XLC / XLRE** | Absent from clustering and signal generation before their data start dates. No backfill applied. |

## Common Traps Checked (Class Notes Section 8.6.3)

| Trap | Status |
|---|---|
| **Look-ahead bias** | Guarded: strict `date >= TRAIN_START` filter; Kalman runs sequentially; clustering uses only past features |
| **Survivorship bias** | Not an issue: all current ETFs existed throughout; XLC/XLRE excluded when absent |
| **Storytelling bias** | Mental model written before code; strategy choice documented in ContextV1.md |
| **Data mining / snooping** | One signal definition; parameter sensitivity shows stable region not cliff-like optimum |
| **Transaction costs** | Noted as exclusion; turnover computed and reported |
| **Outliers** | Omega ratio and % win metrics guard against outlier-driven results |

## Strategy Limitations

**Mean-reversion assumption**: During structural breaks (2008 XLF; 2020 XLE), deviations may persist rather than revert, producing sustained drawdowns. The Kalman filter's slow state adaptation helps but cannot distinguish noise from structural change.

**Cluster instability**: Monthly re-estimation helps but a sector undergoing genuine regime change may be misassigned for several rebalance periods.

**Small universe**: 10–11 ETFs with k=3 gives ~3–4 members per cluster. Leave-one-out benchmarks are sensitive to any single member's idiosyncratic behavior.

**Out-of-sample horizon**: Testing period (Jan 2025 – Mar 2026) is only 15 months — insufficient for conclusive out-of-sample judgment.

## Mental Model Alignment Verification

| Code Component | Mental Model Element | Aligned |
|---|---|---|
| `cluster_at_date()` with `hclust()` | Clustering is the primary signal engine | ✓ |
| `standardize_features()` via `recipes` | Tidymodels preprocessing workflow | ✓ |
| Elbow method + silhouette validation | Validate k, don't guess | ✓ |
| K-Means + Hierarchical comparison | Agreement = robust structure | ✓ |
| Leave-one-out peer mean | Fair value is cluster-relative, not global | ✓ |
| `kalman_local_level()` | Kalman refines cluster-relative signal | ✓ |
| z-score threshold `SIGNAL_THRESHOLD` | Margin of safety | ✓ |
| `MAX_LONGS` / `MAX_SHORTS` | Opportunity cost — strongest signals only | ✓ |
| Month-end only rebalance | Monthly discipline | ✓ |
| Strict train/test + leakage check | Inversion — guard against look-ahead | ✓ |
| `table.Drawdowns()`, `chart.Correlation()` | Full tradeStats risk assessment | ✓ |
| Z-score heatmap + risk appetite filter | Risk-reward framework | ✓ |
